---
title: "RPTEC_EG"
author: "Sem"
date: "2025-05-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
```

```{r}
data_dir <- "C:/Users/semde/Universiteit Leiden/BOO 2025 - BOO CDS Giulia team - BOO CDS Giulia team/Students/Sem/Data/RPTEC_TXG-MAPr"
expression_long   <- readRDS(file.path(data_dir, "expression_long.rds"))
module_definition        <- readRDS(file.path(data_dir, "module_definition.rds"))
eg_score          <- readRDS(file.path(data_dir, "eg_score.rds"))
module_annotation <- readRDS(file.path(data_dir, "module_annotation.rds"))
```



# Distribution of Log2FC and adjusted p-value

## Log2FC

```{r}
ggplot(expression_long, aes(x = log2fc)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "purple", alpha = 0.6) +
  geom_density(color = "darkblue") +
  theme_minimal() + labs(title = "Distribution of Log2FC in RPTEC dataset")
```

```{r}
sum(abs(expression_long$log2fc) > 1) # total log2fc > 1 ("biologically significant" fold change)
```


## p-adjusted

```{r}
ggplot(expression_long, aes(x = padj)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "skyblue", alpha = 0.6) +
  geom_density(color = "black") +
  theme_minimal() +
  labs(title = "Distribution of adjusted p-values in RPTEC dataset")
```

```{r}
sum(is.na(expression_long$padj)) # Na values in padj
```
Something did not went well with the padj values. Will adjust them myself with same method as pHH according to the TXG-MAPr paper (Benjamini Hochberg)

```{r}
expression_long$padj <- p.adjust(expression_long$pvalue, method = "BH")
```

```{r}
sum(is.na(expression_long$padj))
```

```{r}
sum(is.na(expression_long$padj) & is.na(expression_long$log2fc))
```

Still a lot of NA values but there is log2FC data, because these p-values are not significant (source: Giulia). 

```{r}
# Total number of NA genes per condition
expression_long %>%
  group_by(sample_id) %>%
  summarise(n_na = sum(is.na(pvalue))) %>%
  arrange(desc(n_na))
```

```{r}
# Total NA values for the modules with eg_score
sum(is.na(eg_score$eg_score))
```
ALl the modules still have eg_scores, even if there are missing values for padj and log2fc.

## Significant genes

Total statistically significant genes (padj)

```{r}
sum(expression_long$padj <  0.05, na.rm = TRUE)
```

## Both Log2FC and padj significant

total biologically significant genes (log2fc > 1 and padj < 0.05)

```{r}
sum(abs(expression_long$log2fc) > 1 & expression_long$padj < 0.05, na.rm = TRUE)
```

# Gene Response Rate

The proposed Gene Response Rate reflects the total of significant genes (abs(Log2FC) > 1.0 & padj < 0.05) in a module with size n. This will be plotted against the eigengene score to assess this scoring method.

$$
GRR = \frac{ \displaystyle \sum_{i=1}^{n} \left( |\log_2 \mathrm{FC}_i| > 1.0 \ \land \ p_{\mathrm{adj},i} < 0.05 \right) }{ n }
$$

## Merge module_definition with expression_long

**A lot of NA values, can not use module size directly**. Therefore, the module size will be calculated as the number of entries after removal of NA values. 

```{r}
expression_long_filtered <- expression_long %>%
  filter(!is.na(padj), !is.na(log2fc)) # Remove NA values for padj and log2FC
```

```{r}
sum(is.na(expression_long_filtered$padj) & !is.na(expression_long_filtered$log2fc))
```
No NA values left

```{r}
# Add module information, using filtered data
expression_with_module <- expression_long_filtered %>%
  select(entrez_id, sample_id, log2fc, padj) %>%
  inner_join(
    module_definition %>% select(entrez_id, module_number, module_size),
    by = "entrez_id"
  ) %>%
  filter(module_number != 0)  # Remove irrelevant module 0
```

## Calculate significant genes and total per module

Determine what genes are siginificant (log2fc > 1.0 and padj < 0.05)

```{r}
# Calculate significant genes
expression_with_module <- expression_with_module %>%
  mutate(significant = abs(log2fc) > 1 & padj < 0.05)


# Count significant genes per module
sig_count <- expression_with_module %>%
  group_by(module_number, sample_id) %>%
  summarise(sig_genes = sum(significant), .groups = "drop")
```

## Add the module_size to sig_count and calculate GRR

Divide the total significant genes by the module size

```{r}
sig_count <- sig_count %>%
  left_join(
    expression_with_module %>%
      distinct(module_number, module_size),
    by = "module_number"
  )
```

```{r}
sig_count <- sig_count %>%
  mutate(GRR = sig_genes / module_size)
```

## Join GRR with EG score per sample 

```{r}
GRR_vs_eg_score <- sig_count %>%
  inner_join(eg_score, by = c("module_number", "sample_id"))
```

## Density plots of GRR and EG score

```{r}
ggplot(GRR_vs_eg_score, aes(x = GRR)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "green", alpha = 0.6) +
  geom_density(color = "black") +
  theme_minimal() + labs(title = "Distribution of Gene Response Rate in RPTEC dataset")
```
Like expected, the GRR shows a logaritmic distribution due to the count data nature.

```{r}
ggplot(GRR_vs_eg_score, aes(x = eg_score)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "red", alpha = 0.6) +
  geom_density(color = "black") +
  theme_minimal() + labs(title = "Distribution of Eigengene Score in RPTEC dataset")
```
Eigengene shows a normal distribution, as expected due to the PCA use.

```{r}
ggplot(GRR_vs_eg_score, aes(x = abs_eg_score)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "red", alpha = 0.6) +
  geom_density(color = "black") +
  theme_minimal() + labs(title = "Distribution of Absolute Eigengene Score in RPTEC dataset")
```

## Visualize GRR vs EG score with a linear fit

```{r}
ggplot(GRR_vs_eg_score, aes(x = abs_eg_score, y = GRR)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  theme_minimal() +
  labs(
    title = "Gene Response Rate vs Eigengene Score in RPTEC",
    x = "Absolute Eigengene Score (EG)",
    y = "Gene Response Rate (GRR)"
  )
```

The linear fit does not follow the trend of the GRR vs EG, as expected due to the logarithmic distribution of the GRR. Next we will try to fit a different, logarthmic fit. To detect significant outliers.

## Binomial Generalized Linear Model (GLM)

```{r}
glm_fit <- glm(
  cbind(sig_genes, module_size - sig_genes) ~ abs_eg_score,
  family = binomial(link = "logit"),
  data = GRR_vs_eg_score
)
```

```{r}
# Make predictions and residuals
glm_predictions <- predict(glm_fit, type = "response")
glm_dev_resid <- residuals(glm_fit, type = "deviance")

# Combine into a new dataframe
GRR_glm_fit_df <- GRR_vs_eg_score %>%
  mutate(
    fitted_prob = glm_predictions,
    fitted_sig_genes = fitted_prob * module_size,
    residual = sig_genes - fitted_sig_genes,
    dev_resid = glm_dev_resid
  )
```

```{r}
ggplot(GRR_glm_fit_df, aes(x = abs_eg_score, y = GRR)) +
  geom_point(alpha = 0.5, size = 0.8) +
  geom_line(aes(y = fitted_sig_genes / module_size), color = "blue", size = 1) +
  theme_minimal() +
  labs(
    title = "Binomial GLM fit: GRR vs Eigengene Score in RPTEC",
    x = "Eigengene Score",
    y = "Gene Response Rate (GRR)"
  )
```

```{r}
summary(glm_fit)
```

```{r}
deviance(glm_fit) / df.residual(glm_fit)
```

```{r}
1 - (glm_fit$deviance / glm_fit$null.deviance)
```

### significant outliers 

For each module and sample, we compare the observed number of significant genes to the expected number based on the binomial GLM, and calculate a two-sided z-score and p-value using the binomial standard error. By doing this, the significant outlying modules from the glm fit are identified.

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(
    se_fit = sqrt(module_size * fitted_prob * (1 - fitted_prob)),
    z_score = (sig_genes - fitted_sig_genes) / se_fit,
    p_value = 2 * pnorm(-abs(z_score))  # two-sided test
  )
```

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(outlier = p_value < 0.05)
```

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(
    outlier_direction = case_when(
      outlier & z_score > 0 ~ "Above fit",
      outlier & z_score < 0 ~ "Below fit",
      TRUE ~ "Not significant"
    )
  )
```

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(module = paste0("hRPTECTERT1_", module_number)) %>%
  left_join(module_annotation, by = "module")
```

```{r}
sum(GRR_glm_fit_df$outlier, na.rm = TRUE)
```


```{r}
ggplot(GRR_glm_fit_df, aes(x = abs_eg_score, y = GRR)) +
  geom_point(aes(color = outlier), alpha = 0.6, size = 0.9) +
  geom_line(aes(y = fitted_sig_genes / module_size), color = "blue", size = 1) +
  scale_color_manual(values = c("black", "red")) +
  theme_minimal() +
  labs(
    title = "GRR vs Eigengene Score with Significant Outliers in RPTEC",
    x = "Eigengene Score",
    y = "Gene Response Rate (GRR)"
  )
```

## Modules and conditions with most significant outliers'

Based on the binomial GLM fit, we will now find the modules and conditions with the most significant outliers

```{r}
significant_outliers_filtered <- GRR_glm_fit_df %>%
  filter(outlier)
```

```{r}
outliers_per_module <- significant_outliers_filtered %>%
  count(module_number, sort = TRUE)
```

```{r}
outliers_per_sample <- significant_outliers_filtered %>%
  count(sample_id, sort = TRUE)
```

```{r}
head(outliers_per_module, 10) # Top modules with outliers
head(outliers_per_sample, 10) # Top cconditions with outliers
```
### Outliers per module: total, below and above the fit with human annotation

Here, we will visualize the modules with the most outliers. Total, below and above the Binomial GLM fit.

#### Total outliers from the GLM fit

```{r}
threshold <- 300  # Same threshold

GRR_glm_fit_df %>%
  filter(outlier) %>%
  count(module, annotation, sort = TRUE) %>%
  slice_max(n, n = 20, with_ties = FALSE) %>%
  mutate(
    max_n = max(n),
    hjust_val = ifelse(n < threshold, 0, 1.1),
    label_pos = ifelse(n < threshold, n + 2, n - 2)
  ) %>%
  ggplot(aes(x = reorder(module, n))) +
  geom_col(aes(y = max_n), fill = "grey97", width = 0.8) +
  geom_col(aes(y = n), fill = "lavender", width = 0.8) +
  geom_text(
    aes(y = label_pos, label = annotation, hjust = hjust_val),
    size = 1.8, color = "black"
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(
    limits = c(0, max(GRR_glm_fit_df %>%
      filter(outlier) %>%
      count(module) %>%
      slice_max(n, n = 20, with_ties = FALSE) %>%
      pull(n)) + 10),
    expand = c(0, 0)
  ) +
  theme_minimal() +
  theme(
    plot.margin = margin(5.5, 100, 5.5, 5.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Top 20 modules with TOTAL significant GRR outliers (RPTEC)",
    x = "Module",
    y = "Number of significant outliers"
  )
```

#### Outliers above the GLM fit

```{r}
threshold <- 300  # Threshold for switching annotation position

GRR_glm_fit_df %>%
  filter(outlier_direction == "Above fit") %>%
  count(module, annotation, sort = TRUE) %>%
  slice_max(n, n = 20, with_ties = FALSE) %>%
  mutate(
    max_n = max(n),
    hjust_val = ifelse(n < threshold, 0, 1.1),
    label_pos = ifelse(n < threshold, n + 2, n - 2)
  ) %>%
  ggplot(aes(x = reorder(module, n))) +
  geom_col(aes(y = max_n), fill = "grey97", width = 0.8) +
  geom_col(aes(y = n), fill = "seashell", width = 0.8) +
  geom_text(
    aes(y = label_pos, label = annotation, hjust = hjust_val),
    size = 3, color = "black"
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(
    limits = c(0, max(GRR_glm_fit_df %>%
      filter(outlier_direction == "Above fit") %>%
      count(module) %>%
      slice_max(n, n = 20, with_ties = FALSE) %>%
      pull(n)) + 10),
    expand = c(0, 0)
  ) +
  theme_minimal() +
  theme(
    plot.margin = margin(5.5, 100, 5.5, 5.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Top 20 modules with outliers ABOVE the GLM fit (RPTEC)",
    x = "Module",
    y = "Number of outliers above fit"
  )
```

#### Outliers below the GLM fit

```{r}
threshold <- 400  # Threshold for annotation in plot

GRR_glm_fit_df %>%
  filter(outlier_direction == "Below fit") %>%
  count(module, annotation, sort = TRUE) %>%
  slice_max(n, n = 20, with_ties = FALSE) %>%
  mutate(
    max_n = max(n),
    hjust_val = ifelse(n < threshold, 0, 1.1),       # outside or inside
    label_pos = ifelse(n < threshold, n + 2, n - 2)  # offset label position
  ) %>%
  ggplot(aes(x = reorder(module, n))) +
  
  # Background bar
  geom_col(aes(y = max_n), fill = "grey97", width = 0.8) +
  
  # Foreground bar (actual value)
  geom_col(aes(y = n), fill = "lemonchiffon", width = 0.8) +
  
  # Smart annotation positioning
  geom_text(
    aes(y = label_pos, label = annotation, hjust = hjust_val),
    size = 3, color = "black"
  ) +
  
  coord_flip(clip = "off") +  # Flip axes for horizontal bars

  # Y-axis limits: no exessive white space
  scale_y_continuous(limits = c(0, max(GRR_glm_fit_df %>%
    filter(outlier_direction == "Below fit") %>%
    count(module) %>%
    slice_max(n, n = 20, with_ties = FALSE) %>%
    pull(n)) + 10)) +

  theme_minimal() +
  theme(
    plot.margin = margin(5.5, 100, 5.5, 5.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Top 20 modules with outliers BELOW the GLM fit (RPTEC)",
    x = "Module",
    y = "Number of outliers below fit"
  )
```

### Outliers per sample: above, below and total.

Now, we will assess what conditions have the most outliers. Total, below and above the binomial GLM fit.

```{r}
outliers_per_sample %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(sample_id, n), y = n)) +
  geom_col(fill = "darkblue") +
  coord_flip() +
  theme_minimal() +
  ggtitle("Top 20 conditions with significant GRR outliers from the binomial GLM fit (RPTEC)") +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0),
    plot.margin = margin(20, 20, 10, 40) # top, right, bottom, left
  ) +
  labs(
    x = "Condition (sample_id)",
    y = "Number of outliers"
  )
```

```{r}
GRR_glm_fit_df %>%
  filter(outlier_direction == "Above fit") %>%
  count(sample_id, sort = TRUE) %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(sample_id, n), y = n)) +
  geom_col(fill = "brown") +
  coord_flip() +
  theme_minimal() +
  ggtitle("Top 20 conditions with outliers ABOVE the binomial GLM fit (RPTEC)") +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0),
    plot.margin = margin(20, 20, 10, 40)
  ) +
  labs(
    x = "Condition (sample_id)",
    y = "Number of outliers above fit"
  )
```

```{r}
GRR_glm_fit_df %>%
  filter(outlier_direction == "Below fit") %>%
  count(sample_id, sort = TRUE) %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(sample_id, n), y = n)) +
  geom_col(fill = "slateblue") +
  coord_flip() +
  theme_minimal() +
  ggtitle("Top 20 conditions with outliers BELOW the binomial GLM fit (RPTEC)") +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0),
    plot.margin = margin(20, 20, 10, 40)
  ) +
  labs(
    x = "Condition (sample_id)",
    y = "Number of outliers below fit"
  )
```

# Outliers per number of modules * number of samples: density of outliers for comparison with other datasets and techniques.

To compare this with PHH the number of outliers will be compared relative to the total modules and conditions. Also, to have a more arbitrary number to compare with the other topology-based scoring methods

```{r}
outlier_density <- GRR_glm_fit_df %>%
  filter(outlier, module_number != 0) %>%
  summarise(
    total_outliers = n(),
    total_modules = n_distinct(module_number),
    total_conditions = n_distinct(sample_id),
    density = total_outliers / (total_modules * total_conditions)
  )
```

```{r}
print(outlier_density)
```

