---
title: "RPTEC_EG"
author: "Sem"
date: "2025-05-09"
output:
  html_document:
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
      position: left
    toc_depth: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r}
library(ggplot2)
library(dplyr)
```
# RPTEC Data

```{r}
data_dir <- data_dir <- "C:/Users/semde/Documents/BOO_Scripts/Data/RPTEC_TXG-MAPr"
expression_long   <- readRDS(file.path(data_dir, "expression_long.rds"))
module_definition        <- readRDS(file.path(data_dir, "module_definition.rds"))
eg_score          <- readRDS(file.path(data_dir, "eg_score.rds"))
module_annotation <- readRDS(file.path(data_dir, "module_annotation.rds"))
```

# Introduction

The eigengene scoring method will be evaluated in the Renal Proximal Tubule Epithelial Cell (RPTEC) dataset from the Leiden Academic Center for Drug Research (LACDR). It will be evaluated on the Gene Response Rate and  Weighted Gene Response Rate (wGRR) metric, which is proposed as an evaluation metric for the topological activation of every unique combination of condition and module. The GRR and wGRR are presented as the following formulas:

$$
\mathrm{GRR}(M_j) = \frac{ \displaystyle \sum_{i=1}^{n_j} \left( |\log_2 \mathrm{FC}_i| > 1.0 \,\wedge\, p_{\mathrm{adj},i} < 0.05 \right)} { n_j }
$$

$$
\mathrm{wGRR}(M_j) = \frac{ \displaystyle \sum_{i=1}^{n_j} \left( |\log_2 \mathrm{FC}_i| > 1.0 \,\wedge\, p_{\mathrm{adj},i} < 0.05 \right) \cdot \mathrm{median(TOM)}_i }{ \sum_{i=1}^{n_j} \mathrm{median(TOM)}_i }
$$
The GRR represents the proportion of significantly expressed genes in a module (abs log2fc > 1.0 and padj < 0.05) and the wGRR weighs this proportion by the median(TOM) of a gene within a module.

# Distribution of Log2FC and adjusted p-value

First, the distributions of the Log2fc and adj p-value in the RPTEC dataset will be visualized

## Log2FC

```{r}
ggplot(expression_long, aes(x = log2fc)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "purple", alpha = 0.6) +
  geom_density(color = "darkblue") +
  theme_minimal() + labs(title = "Distribution of Log2FC in RPTEC dataset")
```
The log2fc shows a normal distribution in the RPTEC dataset

```{r}
sum(abs(expression_long$log2fc) > 1) # total log2fc > 1 ("biologically significant" fold change)
```
There are a total of 653992 genes with a significant log2fc.

## p-adjusted

```{r}
ggplot(expression_long, aes(x = padj)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "skyblue", alpha = 0.6) +
  geom_density(color = "black") +
  theme_minimal() +
  labs(title = "Distribution of adjusted p-values in RPTEC dataset")
```

The padj values are not normally distributed, as expected due to it being a pvalue.

```{r}
sum(is.na(expression_long$padj)) # Na values in padj
```
The pvalues are not adjusted yet. Will adjust them myself with same method as pHH according to the TXG-MAPr paper (Benjamini Hochberg)

```{r}
expression_long$padj <- p.adjust(expression_long$pvalue, method = "BH")
```

```{r}
sum(is.na(expression_long$padj))
```
Now there are still some NA values in the padj column. According to Giulia, this is because they are not significant.

```{r}
sum(is.na(expression_long$padj) & is.na(expression_long$log2fc))
```
To prove Giulias point. There are no log2fc NA values AND adj pvalues that are NA. 

```{r}
# Total number of NA genes per condition
expression_long %>%
  group_by(sample_id) %>%
  summarise(n_na = sum(is.na(pvalue))) %>%
  arrange(desc(n_na))
```

```{r}
# Total NA values for the modules with eg_score
sum(is.na(eg_score$eg_score))
```
ALl the modules still have eg_scores, even if there are missing values for padj and log2fc. So we will proceed with the analysis

## Significant genes

Total statistically significant genes (padj)

```{r}
sum(expression_long$padj <  0.05, na.rm = TRUE)
```
There are 232911 genes in the RPTEC dataset that have significant padj values.

## Both Log2FC and padj significant

total biologically significant genes (log2fc > 1 and padj < 0.05)

```{r}
sum(abs(expression_long$log2fc) > 1 & expression_long$padj < 0.05, na.rm = TRUE)
```
There are 184290 genes that are biologically and statistically significant in the RPTEC dataset.

## Gene Response Rate

**The GRR calculation here is not relevant anymore, because the calculation was changed and calculated in the hRPTEC_DRW.Rmd file. But the chunks are still included. Because some parts are used in the fitting **


## Merge module_definition with expression_long


```{r}
expression_long_filtered <- expression_long %>%
  filter(!is.na(padj), !is.na(log2fc)) # Remove NA values
```

```{r}
sum(is.na(expression_long_filtered$padj) & !is.na(expression_long_filtered$log2fc))
```

No NA values left

```{r}
# Add module information, using filtered data
expression_with_module <- expression_long_filtered %>%
  select(entrez_id, sample_id, log2fc, padj) %>%
  inner_join(
    module_definition %>% select(entrez_id, module_number, module_size),
    by = "entrez_id"
  ) %>%
  filter(module_number != 0)  # Remove irrelevant module 0
```

## Calculate significant genes and total per module

Determine what genes are siginificant (log2fc > 1.0 and padj < 0.05)

```{r}
# Calculate significant genes
expression_with_module <- expression_with_module %>%
  mutate(significant = abs(log2fc) > 1 & padj < 0.05)


# Count significant genes per module
sig_count <- expression_with_module %>%
  group_by(module_number, sample_id) %>%
  summarise(sig_genes = sum(significant), .groups = "drop")
```

## Add the module_size to sig_count and calculate GRR

Divide the total significant genes by the module size

```{r}
sig_count <- sig_count %>%
  left_join(
    expression_with_module %>%
      distinct(module_number, module_size),
    by = "module_number"
  )
```

```{r}
sig_count <- sig_count %>%
  mutate(GRR = sig_genes / module_size)
```

## Join GRR with EG score per sample 

```{r}
GRR_vs_eg_score <- sig_count %>%
  inner_join(eg_score, by = c("module_number", "sample_id"))
```

```{r}
#saveRDS(GRR_vs_eg_score, file = "GRR_vs_eg_score_RPTEC.rds")
```


# wGRR and (correct) GRR

It was later decided that the wGRR metric was more representative of topological activation than the GRR metric. This metric was calculated in the hRPTEC_DRW.rmd file and loaded in to visualize it againt the abs_eg_score. The wGRR improves the GRR metric by weighting the genes by their cor_egs, which is used as a measure for their topological importance:

```{r}
GRR_vs_DRW   <- readRDS(file.path(data_dir, "GRR_vs_DRW_cor_eg_RPTEC.rds"))
```

```{r}
cols_from_DRW <- GRR_vs_DRW %>%
  select(sample_id, module_number, wGRR, GRR)
```

```{r}
GRR_vs_eg_score <- GRR_vs_eg_score %>%
  select(-GRR) %>% 
  left_join(cols_from_DRW, by = c("sample_id", "module_number")) # Joining the wGRR with the eg_scores
```

## Density plots of GRR, wGRR and DRW score

```{r}
ggplot(GRR_vs_eg_score, aes(x = abs(GRR))) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "green", alpha = 0.6) +
  geom_density(color = "black") +
  theme_minimal() + labs(title = "Distribution of Gene Response Rate (GRR) in RPTEC dataset")
```

```{r}
ggplot(GRR_vs_eg_score, aes(x = abs(wGRR))) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "blue", alpha = 0.6) +
  geom_density(color = "black") +
  theme_minimal() + labs(title = "Distribution of Weighted Gene Response Rate (wGRR) in RPTEC dataset")
```

Both abs(wGRR) and abs(GRR) log distributed, as expected due to abs transformation

# Fitting the wGRR and GRR vs EG score using binomial GLMs

## Visualize GRR and wGRR vs DRW score with a linear fit

```{r}
ggplot(GRR_vs_eg_score, aes(x = eg_score, y = abs(GRR))) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  theme_minimal() +
  labs(
    title = "Abs Gene Response Rate (GRR) vs DRW Score in RPTEC",
    x = "Eigengene Score (EG)",
    y = "Abs Gene Response Rate (GRR)"
  )
```

```{r}
ggplot(GRR_vs_eg_score, aes(x = eg_score, y = abs(wGRR))) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  theme_minimal() +
  labs(
    title = "Abs Weighted Gene Response Rate (wGRR) vs DRW Score in RPTEC",
    x = "Eigengene Score (EG)",
    y = "Abs Weighted Gene Response Rate (wGRR)"
  )
```

The linear fit does not fit the data well visually, so a binomial glm will be fitted and outliers will be calculated from this fit.

## Binomial Generalized Linear Model (GLM): GRR vs DRW

```{r}
glm_data <- GRR_vs_eg_score %>%
  filter(!is.na(GRR), !is.na(eg_score)) # This filtering was performed as the function was first tested on a subset, but the filtering is irrelevant for the whole dataset as there are no NAs anymore
```


```{r}
glm_fit <- glm(
  abs(GRR) ~ abs(eg_score), # absolute because of binomial fit
  data = glm_data,
  family = quasibinomial(link = "logit") # quasibinomial as the GRR is a proportion
)
```

```{r}
# Make predictions and residuals
glm_predictions <- predict(glm_fit, type = "response")
glm_dev_resid <- residuals(glm_fit, type = "deviance")

# Combine into a new dataframe
GRR_glm_fit_df <- glm_data %>%
  mutate(
    fitted_GRR = glm_predictions,
    GRR_resid = abs(GRR) - fitted_GRR, # Calculating the resid for outlier calculation
    dev_resid = glm_dev_resid
  )
```

```{r}
ggplot(GRR_glm_fit_df, aes(x = eg_score, y = abs(GRR))) +
  geom_point(alpha = 0.5, size = 0.8) +
  geom_line(aes(y = fitted_GRR), color = "blue", size = 1) +
  theme_minimal() +
  labs(
    title = "Binomial GLM fit: Abs GRR vs DRW in RPTEC",
    x = "Eigengene Score (EG)",
    y = "Absolute Gene Response Rate (GRR)"
  )
```

Binomial fit seems a good fit visually. Lots of weird values with high EG scores and low GRR

```{r}
summary(glm_fit)
```
Statistically it is also a good fit.

### significant outliers 

For each module and sample, we compare the observed number of significant genes to the expected number based on the binomial GLM, and calculate a two-sided z-score and p-value using the binomial standard error. By doing this, the significant outlying modules from the glm fit are identified.

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(
    se_fit = sqrt(fitted_GRR * (1 - fitted_GRR) / module_size),
    z_score = GRR_resid / se_fit,
    p_value = 2 * pnorm(-abs(z_score))  # two-sided test
  )
```

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(
    outlier = p_value < 0.05,
    outlier_direction = case_when(
      outlier & z_score > 0 ~ "Above fit",
      outlier & z_score < 0 ~ "Below fit",
      TRUE ~ "Not significant"
    )
  )
```

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(module = paste0("hRPTECTERT1_", module_number)) %>%
  left_join(module_annotation, by = "module")
```

```{r}
sum(GRR_glm_fit_df$outlier, na.rm = TRUE)
```
There are 8862 outliers found from the GRR fit


```{r}
total_modules_all <- GRR_glm_fit_df %>%
  filter(module_number != 0) %>%
  summarise(n = n_distinct(module_number)) %>%
  pull(n) # for normalization
```


```{r}
total_conditions_all <- GRR_glm_fit_df %>%
  filter(module_number != 0) %>%
  summarise(n = n_distinct(sample_id)) %>%
  pull(n) # for normalization
```


```{r}
# Make a summarizing df for comparing with other datasets and scoring algorithms
outlier_density <- GRR_glm_fit_df %>%
  filter(outlier, module_number != 0) %>%
  group_by(outlier_direction) %>%
  summarise(
    total_outliers = n(),
    .groups = "drop"
  ) %>%
  bind_rows(
    tibble(
      outlier_direction = "Total",
      total_outliers = sum(.$total_outliers)
    )
  ) %>%
  mutate(
    total_modules = total_modules_all,
    total_conditions = total_conditions_all,
    density = total_outliers / (total_modules * total_conditions)
  )
```

```{r}
print(outlier_density)
```

Above shows the density of the outliers. n_outliers / (n_conditions * n_modules)

```{r}
ggplot(GRR_glm_fit_df, aes(x = abs(eg_score), y = abs(GRR))) +
  geom_point(aes(color = outlier), alpha = 0.6, size = 0.9) +
  geom_line(aes(y = fitted_GRR), color = "blue", size = 1) +
  scale_color_manual(values = c("black", "red")) +
  theme_minimal() +
  labs(
    title = "abs(GRR) vs abs(EG) Score with Significant Outliers in RPTEC",
    x = "Absolute Eigengene Score (EG)",
    y = "Absolute Gene Response Rate (GRR)"
  )
```

```{r}
ggplot(GRR_glm_fit_df, aes(x = eg_score, y = abs(GRR))) +
  geom_point(aes(color = outlier), alpha = 0.6, size = 0.9) +
  geom_line(aes(y = fitted_GRR), color = "blue", size = 1) +
  scale_color_manual(values = c("black", "red")) +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  labs(
    title = "Eigengene (RPTEC)",
    x = "Eigengene Score (EG)",
    y = "Absolute Gene Response Rate (GRR)"
  )
```


## Binomial Generalized Linear Model (GLM): wGRR vs DRW

```{r}
glm_data <- GRR_vs_DRW %>%
  filter(!is.na(wGRR), !is.na(eg_score)) # This filtering was performed as the function was first tested on a subset, but the filtering is irrelevant for the whole dataset as there are no NAs
```


```{r}
glm_fit <- glm(
  abs(wGRR) ~ abs(eg_score),
  data = glm_data,
  family = quasibinomial(link = "logit") # quasibinomial as the wGRR is a proportion multiplied by a continuous variable, the median(TOM)
)
```

```{r}
# Make predictions and residuals
glm_predictions <- predict(glm_fit, type = "response")
glm_dev_resid <- residuals(glm_fit, type = "deviance")

# Combine into a new dataframe
GRR_glm_fit_df <- glm_data %>%
  mutate(
    fitted_wGRR = glm_predictions,
    wGRR_resid = abs(wGRR) - fitted_wGRR,
    dev_resid = glm_dev_resid
  )
```

```{r}
ggplot(GRR_glm_fit_df, aes(x = eg_score, y = abs(wGRR))) +
  geom_point(alpha = 0.5, size = 0.8) +
  geom_line(aes(y = fitted_wGRR), color = "blue", size = 1) +
  theme_minimal() +
  labs(
    title = "Binomial GLM fit: wGRR vs DRW in RPTEC",
    x = "Eigengene Score (EG)",
    y = "Weighted Gene Response Rate (wGRR)"
  )
```

Some weird outlying values with0 EG scores and high wGRR and high EG scores with low wGRR. The fit seems good visually.

```{r}
summary(glm_fit)
```
Statistically the binomial glm fit is good as well.

### significant outliers 

For each module and sample, we compare the observed number of significant genes to the expected number based on the binomial GLM, and calculate a two-sided z-score and p-value using the binomial standard error. By doing this, the significant outlying modules from the glm fit are identified.

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(
    se_fit = sqrt(fitted_wGRR * (1 - fitted_wGRR) / module_size),
    z_score = wGRR_resid / se_fit,
    p_value = 2 * pnorm(-abs(z_score))  # two-sided test
  )
```

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(
    outlier = p_value < 0.05,
    outlier_direction = case_when(
      outlier & z_score > 0 ~ "Above fit",
      outlier & z_score < 0 ~ "Below fit",
      TRUE ~ "Not significant"
    )
  )
```

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(module = paste0("hRPTECTERT1_", module_number)) %>%
  left_join(module_annotation, by = "module")
```

```{r}
sum(GRR_glm_fit_df$outlier, na.rm = TRUE)
```
There are 9675 outliers found from the wGRR binomial fit in the RPTEC dataset.

```{r}
ggplot(GRR_glm_fit_df, aes(x = abs(eg_score), y = abs(wGRR))) +
  geom_point(aes(color = outlier), alpha = 0.6, size = 0.9) +
  geom_line(aes(y = fitted_wGRR), color = "blue", size = 1) +
  scale_color_manual(values = c("black", "red")) + 
  theme_minimal() +
  labs(
    title = "abs(wGRR) vs abs(EG) Score with Significant Outliers in RPTEC",
    x = "Absolute Eigengene Score (EG)",
    y = "Absolute Weighted Gene Response Rate (wGRR)"
  )
```

```{r}
ggplot(GRR_glm_fit_df, aes(x = eg_score, y = abs(wGRR))) +
  geom_point(aes(color = outlier), alpha = 0.6, size = 0.9) +
  geom_line(aes(y = fitted_wGRR), color = "blue", size = 1) +
  scale_color_manual(values = c("black", "red")) +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  labs(
    title = "Eigengene (RPTEC)",
    x = "Eigengene Score (EG)",
    y = "Absolute Weighted Gene Response Rate (wGRR)"
  )
```

# Modules and conditions with most significant outliers'

Based on the binomial GLM fit, we will now find the modules and conditions with the most significant outliers

```{r}
significant_outliers_filtered <- GRR_glm_fit_df %>%
  filter(outlier)
```

```{r}
outliers_per_module <- significant_outliers_filtered %>%
  count(module_number, sort = TRUE)
```

```{r}
outliers_per_sample <- significant_outliers_filtered %>%
  count(sample_id, sort = TRUE)
```

```{r}
head(outliers_per_module, 10) # Top modules with outliers
head(outliers_per_sample, 10) # Top cconditions with outliers
```

### Outliers per module: total, below and above the fit with human annotation

Here, we will visualize the modules with the most outliers and its human annotation. Total, below and above the Binomial GLM fit.

#### Total outliers from the GLM fit

```{r}
threshold <- 300  # threshold for annotation

GRR_glm_fit_df %>%
  filter(outlier) %>%
  count(module, annotation, sort = TRUE) %>%
  slice_max(n, n = 20, with_ties = FALSE) %>%
  mutate(
    max_n = max(n),
    hjust_val = ifelse(n < threshold, 0, 1.1),
    label_pos = ifelse(n < threshold, n + 2, n - 2)
  ) %>%
  ggplot(aes(x = reorder(module, n))) +
  geom_col(aes(y = max_n), fill = "grey97", width = 0.8) +
  geom_col(aes(y = n), fill = "lavender", width = 0.8) +
  geom_text(
    aes(y = label_pos, label = annotation, hjust = hjust_val),
    size = 1.8, color = "black"
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(
    limits = c(0, max(GRR_glm_fit_df %>%
      filter(outlier) %>%
      count(module) %>%
      slice_max(n, n = 20, with_ties = FALSE) %>%
      pull(n)) + 10),
    expand = c(0, 0)
  ) +
  theme_minimal() +
  theme(
    plot.margin = margin(5.5, 100, 5.5, 5.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Top 20 modules with TOTAL significant GRR outliers (RPTEC) (EG)",
    x = "Module",
    y = "Number of significant outliers"
  )
```

#### Outliers above the GLM fit

```{r}
threshold <- 300  # Threshold for switching annotation position

GRR_glm_fit_df %>%
  filter(outlier_direction == "Above fit") %>%
  count(module, annotation, sort = TRUE) %>%
  slice_max(n, n = 20, with_ties = FALSE) %>%
  mutate(
    max_n = max(n),
    hjust_val = ifelse(n < threshold, 0, 1.1),
    label_pos = ifelse(n < threshold, n + 2, n - 2)
  ) %>%
  ggplot(aes(x = reorder(module, n))) +
  geom_col(aes(y = max_n), fill = "grey97", width = 0.8) +
  geom_col(aes(y = n), fill = "seashell", width = 0.8) +
  geom_text(
    aes(y = label_pos, label = annotation, hjust = hjust_val),
    size = 3, color = "black"
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(
    limits = c(0, max(GRR_glm_fit_df %>%
      filter(outlier_direction == "Above fit") %>%
      count(module) %>%
      slice_max(n, n = 20, with_ties = FALSE) %>%
      pull(n)) + 10),
    expand = c(0, 0)
  ) +
  theme_minimal() +
  theme(
    plot.margin = margin(5.5, 100, 5.5, 5.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Top 20 modules with outliers ABOVE the GLM fit (RPTEC) (EG)",
    x = "Module",
    y = "Number of outliers above fit"
  )
```

#### Outliers below the GLM fit

```{r}
threshold <- 400  # Threshold for annotation in plot

GRR_glm_fit_df %>%
  filter(outlier_direction == "Below fit") %>%
  count(module, annotation, sort = TRUE) %>%
  slice_max(n, n = 20, with_ties = FALSE) %>%
  mutate(
    max_n = max(n),
    hjust_val = ifelse(n < threshold, 0, 1.1),       # outside or inside
    label_pos = ifelse(n < threshold, n + 2, n - 2)  # offset label position
  ) %>%
  ggplot(aes(x = reorder(module, n))) +
  
  # Background bar
  geom_col(aes(y = max_n), fill = "grey97", width = 0.8) +
  
  # Foreground bar (actual value)
  geom_col(aes(y = n), fill = "lemonchiffon", width = 0.8) +
  
  # Smart annotation positioning
  geom_text(
    aes(y = label_pos, label = annotation, hjust = hjust_val),
    size = 3, color = "black"
  ) +
  
  coord_flip(clip = "off") +  # Flip axes for horizontal bars

  # Y-axis limits: no exessive white space
  scale_y_continuous(limits = c(0, max(GRR_glm_fit_df %>%
    filter(outlier_direction == "Below fit") %>%
    count(module) %>%
    slice_max(n, n = 20, with_ties = FALSE) %>%
    pull(n)) + 10)) +

  theme_minimal() +
  theme(
    plot.margin = margin(5.5, 100, 5.5, 5.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Top 20 modules with outliers BELOW the GLM fit (RPTEC) (EG)",
    x = "Module",
    y = "Number of outliers below fit"
  )
```

### Outliers per sample: above, below and total.

Now, we will assess what conditions have the most outliers. Total, below and above the binomial GLM fit.

```{r}
outliers_per_sample %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(sample_id, n), y = n)) +
  geom_col(fill = "darkblue") +
  coord_flip() +
  theme_minimal() +
  ggtitle("Top 20 conditions with significant GRR outliers from the binomial GLM fit (RPTEC) (EG)") +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0),
    plot.margin = margin(20, 20, 10, 40) # top, right, bottom, left
  ) +
  labs(
    x = "Condition (sample_id)",
    y = "Number of outliers"
  )
```

```{r}
GRR_glm_fit_df %>%
  filter(outlier_direction == "Above fit") %>%
  count(sample_id, sort = TRUE) %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(sample_id, n), y = n)) +
  geom_col(fill = "brown") +
  coord_flip() +
  theme_minimal() +
  ggtitle("Top 20 conditions with outliers ABOVE the binomial GLM fit (RPTEC) (EG)") +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0),
    plot.margin = margin(20, 20, 10, 40)
  ) +
  labs(
    x = "Condition (sample_id)",
    y = "Number of outliers above fit"
  )
```

```{r}
GRR_glm_fit_df %>%
  filter(outlier_direction == "Below fit") %>%
  count(sample_id, sort = TRUE) %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(sample_id, n), y = n)) +
  geom_col(fill = "slateblue") +
  coord_flip() +
  theme_minimal() +
  ggtitle("Top 20 conditions with outliers BELOW the binomial GLM fit (RPTEC) (EG)") +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0),
    plot.margin = margin(20, 20, 10, 40)
  ) +
  labs(
    x = "Condition (sample_id)",
    y = "Number of outliers below fit"
  )
```

# Conclusion

The Eigengene score shows improveable performance on the GRR and wGRR metrics in the RPTEC dataset. There are lots of low EG scores with high GRR and wGRR points and there are some high EG scores with low GRR and wGRR points. The aim is to improve this using the Weighted Directed Random Walk algorithm, which will be assessed in the hRPTEC_wDRW.Rmd file.
