---
title: "CTD Validation"
author: "Sem"
date: "2025-05-22"
output:
  html_document:
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
      position: left
    toc_depth: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(dplyr)
library(stringr)
library(purrr)
library(ggplot2)
```

# Introduction

In this .Rmd file, the top-k scoring combination of modules and conditions are identified for different scoring algorithms and conditions. For each compound, the Reactome-based enrichment of these modules is compared to the Reactome compound enrichment from the Comparative Toxicogenomics Database(CTD) database. Only Reactome pathways were used for this comparison, as it was the only source shared between the module enrichment results and the CTD data. This analysis serves to evaluate the biological relevance of the proposed module scoring methods. It is assessed if there are matches between the databases (0= no match, 1 = 1 or more match). Then the accuracy for the top-k combinations of modules and conditions per scoring method and dataset will be computed to evaluate if DRW based scoring provide more biologically accurate results

## CTD data

```{r}
data_dir <- "C:/Users/semde/Documents/BOO_Scripts/Data/Evaluation"
CTD_Pathways <- read_csv(
  file.path(data_dir, "CTD_Pathways_enriched.csv"),
  skip = 27,
  show_col_types = FALSE
)
```
First row NA warning, will be fixed later on.

## PHH data

```{r}
data_dir <- "C:/Users/semde/Documents/BOO_Scripts/Data/PHH_TXG-MAPr"
module_enrichment_PHH   <- readRDS(file.path(data_dir, "module_enrichment.rds"))
drw_scores_PHH <- readRDS(file.path(data_dir, "drw_scores_PHH_r0.7.rds"))
eg_scores_PHH <- readRDS(file.path(data_dir, "eg_score.rds"))
```

## RPTEC data

```{r}
data_dir <- "C:/Users/semde/Documents/BOO_Scripts/Data/RPTEC_TXG-MAPr"
module_enrichment_RPTEC   <- readRDS(file.path(data_dir, "module_enrichment.rds"))
drw_scores_RPTEC <- readRDS(file.path(data_dir, "drw_scores_RPTEC_r0.7.rds"))
eg_scores_RPTEC <- readRDS(file.path(data_dir, "eg_score.rds"))
```

# Preprocessing

## CTD

First row has NA due to weird csv (# in first line)

```{r}
CTD_Pathways_clean <- CTD_Pathways[-1, ]
```

Reactome is the only match between the module enrichment and CTD

```{r}
CTD_Pathways_reactome <- CTD_Pathways_clean %>%
  filter(str_starts(PathwayID, "REACT:")) %>% 
  filter(CorrectedPValue < 0.05)
```

```{r}
glimpse(CTD_Pathways_reactome)
```
Filter the reactome code out of the Pathwayname for matching with the RPTEC dataset

```{r}
CTD_Pathways_reactome <- CTD_Pathways_reactome %>%
  mutate(reactome_id = str_extract(PathwayID, "R-HSA-\\d+"))
```


## PHH

### enrichment df

Only include significantly enriched terms

```{r}
module_enrichment_PHH <- module_enrichment_PHH %>%
  filter(padj < 0.05)
```

```{r}
unique(module_enrichment_PHH$source)
```
Filter for only reactome terms

```{r}
module_enrichment_PHH_reactome <- module_enrichment_PHH %>%
  filter(source == "CPDB:Reactome")
```

get the module_number

```{r}
module_enrichment_PHH_reactome <- module_enrichment_PHH_reactome %>%
  mutate(module_number = str_remove(module, "^hPHH_"),
         module_number = as.integer(module_number))
```


### eg_score df

Strip for module numbers

```{r}
eg_scores_PHH <- eg_scores_PHH %>%
  mutate(chemical = str_extract(sample_id, "(?<=TG_HPHH_SINGLE_)[^_]+")) %>% 
  select(abs_eg_score, chemical, module_number)
```

### drw_score df

```{r}
drw_scores_PHH <- drw_scores_PHH %>%
  mutate(chemical = str_extract(condition_id, "(?<=TG_HPHH_SINGLE_)[^_]+")) %>% 
  mutate(module_number=module_nr) %>%
  select(chemical, drw_zscore, module_number)
  
```


## RPTEC

```{r}
module_enrichment_RPTEC <- module_enrichment_RPTEC %>%
  filter(padj < 0.05)
```

```{r}
unique(module_enrichment_RPTEC$source)
```

Filter for only reactome terms

```{r}
module_enrichment_RPTEC_reactome <- module_enrichment_RPTEC %>%
  filter(source == "Reactome_2022")
```

```{r}
module_enrichment_RPTEC_reactome <- module_enrichment_RPTEC_reactome %>%
  mutate(term_id = str_extract(term, "R-HSA-\\d+"))
```

get the module number

```{r}
module_enrichment_RPTEC_reactome <- module_enrichment_RPTEC_reactome %>%
  mutate(module_number = str_remove(module, "^hRPTECTERT1_"),
         module_number = as.integer(module_number))
```


### eg_score df

Strip for module numbers

```{r}
eg_scores_RPTEC <- eg_scores_RPTEC %>%
  mutate(chemical = str_extract(sample_id, "(?<=LU_HRPTECTERT1_SINGLE_)[^_]+")) %>% 
  select(abs_eg_score, chemical, module_number)
```

### drw_score df

```{r}
drw_scores_RPTEC <- drw_scores_RPTEC %>%
  mutate(chemical = str_extract(condition_id, "(?<=LU_HRPTECTERT1_SINGLE_)[^_]+")) %>% 
  mutate(module_number=module_nr) %>%
  select(chemical, drw_zscore, module_number)
  
```

# Seeing if there are matches

To see if this analysis is possible, it was first checked to see if there are matches.

For PHH matching needs to be done on strings, which makes it less good. But there are no reactome codes available. 

```{r}
# 1. Extract the two name‐vectors
ctd_names   <- CTD_Pathways_reactome$PathwayName
phh_names   <- module_enrichment_PHH_reactome$term

# 2. Find exact overlaps
common_names <- intersect(ctd_names, phh_names)

# 3. Subset the enrichment table to only those matches
matches_df_PHH <- module_enrichment_PHH %>%
  filter(term %in% common_names)
```

```{r}
nrow(matches_df_PHH)
```
For RPTEC matching could be done on R-HSA code.

```{r}
ctd_ids   <- CTD_Pathways_reactome$reactome_id
rptec_ids <- module_enrichment_RPTEC_reactome$term_id

# Find common id
common_ids <- intersect(ctd_ids, rptec_ids)

matches_df_RPTEC <- module_enrichment_RPTEC_reactome %>%
  filter(term_id %in% common_ids)
```

```{r}
nrow(matches_df_RPTEC)
```


The analysis seems possible as there are string matches between the CTD database and the PHH TXG-MAPr enrichment. And reactome_id matches between the RPTEC TXG-MAPr enrichment and PHH TXG-MAPr enrichment. It is noteable that there are less matches between the strings. THis could be an indication that the enrichment should be done again. But for now, not enough time. But then again there are also less overal enrichment terms for PHH.

# PHH analysis

## Preprocessing (again)

First, the chemicals from TXG-MAPr have to match the chemical for the CTD database. This was manually done for the top k pathways

### Arranging the CTD

arranging the ctd database for manual mapping of the chemicals.

```{r}
CTD_chemicals_unique <- CTD_Pathways_reactome %>%
  arrange(`# ChemicalName`) %>%
  distinct(`# ChemicalName`, .keep_all = TRUE)
```


### eg_score

```{r}
eg_scores_PHH %>%
  arrange(desc(abs_eg_score)) %>%
  distinct(chemical, module_number, .keep_all = TRUE) %>%
  head(100)
```

```{r}
eg_scores_PHH <- eg_scores_PHH %>%
  mutate(chemical_ctd = case_when(
    chemical == "DOXORUBICIN"     ~ "Doxorubicin",
    chemical == "PAPAVERINE"  ~ "Papaverine",
    chemical == "VALPROICACID"    ~ "Valproic Acid",
    chemical == "CYCLOHEXIMIDE"    ~ "Cycloheximide",
    chemical == "INTERFERONALPHAHUMAN" ~ "Interferon-alpha",
    chemical == "TRANSFORMINGGROWTHFACTORBETA1" ~ "Transforming Growth Factor beta1",
    chemical == "ETHIONINE" ~ "Ethionine",
    chemical == "PHENOBARBITAL" ~ "Phenobarbital",
    chemical == "CAFFEINE" ~ "Caffeine",
    chemical == "BROMOETHYLAMINE" ~ "2-bromoethylamine",
    chemical == "ALLYLALCOHOL" ~ "allyl alcohol",
    chemical == "NEFAZODONE" ~ "nefazodone",
    chemical == "THEOPHYLLINE" ~ "Theophylline",
    chemical == "PHALLOIDIN" ~ "Phalloidine",
    chemical == "OMEPRAZOLE" ~ "Omeprazole",
    chemical == "SULINDAC" ~ "Sulindac",
    chemical == "COLCHICINE" ~ "Colchicine",
    chemical == "AFLATOXINB1" ~ "Aflatoxin B1",
    chemical == "NMETHYLNNITROSOUREA" ~ "Methylnitrosourea",
    chemical == "TNFALPHARAT" ~ "Tumor Necrosis Factors",
    chemical == "TUNICAMYCIN" ~ "Tunicamycin",
    chemical == "ACETAMINOPHEN" ~ "Acetaminophen",
    chemical == "INTERLEUKIN1BETAHUMAN" ~ "Interleukins",
    TRUE ~ chemical  # fallback to original if no match
  ))
```

```{r}
eg_scores_PHH %>%
  arrange(desc(abs_eg_score)) %>%
  distinct(chemical, module_number, .keep_all = TRUE) %>%
  head(100)
```

```{r}
top100_results_PHH_eg <- eg_scores_PHH %>%
  arrange(desc(abs_eg_score)) %>%
  distinct(chemical, module_number, .keep_all = TRUE) %>%
  slice_head(n = 100) %>%
  transmute(chemical_ctd,
            module_nr = module_number,
            abs_eg_score) %>%    # renaming because of joining problems
  rowwise() %>%
  mutate(
    module_terms = list(
      module_enrichment_PHH_reactome %>%
        filter(
          module_number == module_nr,       
          source        == "CPDB:Reactome", # already did this but just to be sure
          padj          <  0.05 # same
        ) %>%
        pull(term) %>%
        unique()
    ),
    ctd_terms = list(
      CTD_Pathways_reactome %>%
        filter(
          `# ChemicalName` == chemical_ctd,
          CorrectedPValue  <  0.05
        ) %>%
        pull(PathwayName) %>%
        unique()
    ),
    overlaps  = list(intersect(module_terms, ctd_terms)),
    n_matches = length(overlaps),
    has_match = as.integer(n_matches > 0)
  ) %>%
  ungroup() %>%
  select(chemical_ctd,
         module_number = module_nr,
         n_matches,
         has_match,
         abs_eg_score)
```


```{r}
print(top100_results_PHH_eg)
```


### drw_score

```{r}
drw_scores_PHH %>%
  arrange(desc(abs(drw_zscore))) %>%
  distinct(chemical, module_number, .keep_all = TRUE) %>%
  head(100)
```


```{r}
drw_scores_PHH <- drw_scores_PHH %>%
  mutate(chemical_ctd = case_when(
    chemical == "OMEPRAZOLE"     ~ "Omeprazole",
    chemical == "ISONIAZID"  ~ "Isoniazid",
    chemical == "2NITROFLUORENE"    ~ "2-nitrofluorene",
    chemical == "HEXACHLOROBENZENE" ~ "Hexachlorobenzene",
    chemical == "SULINDAC" ~ "Sulindac",
    chemical == "ACETAMIDOFLUORENE" ~ "4-acetylaminofluorene",
    chemical == "FLUTAMIDE" ~ "Flutamide",
    chemical == "MEXILETINE" ~ "Mexiletine",
    chemical == "PHALLOIDIN" ~ "Phalloidine",
    chemical == "ALLYLALCOHOL" ~ "allyl alcohol",
    chemical == "DOXORUBICIN" ~ "Doxorubicin",
    chemical == "NIFEDIPINE" ~ "Nifedipine",
    chemical == "TNFALPHARAT" ~ "Tumor Necrosis Factors",
    chemical == "CYCLOHEXIMIDE" ~ "Cycloheximide",
    chemical == "ACETAMINOPHEN" ~ "Acetaminophen",
    chemical == "ETHIONINE" ~ "Ethionine",
    chemical == "INTERLEUKIN1BETAHUMAN" ~ "Interleukins",
    chemical == "COLCHICINE" ~ "Colchicine",
    chemical == "NAPROXEN" ~ "Naproxen",
    chemical == "THEOPHYLLINE" ~ "Theophylline",
    chemical == "VALPROICACID"    ~ "Valproic Acid",
    chemical == "METHAPYRILENE" ~ "Methapyrilene",
    chemical == "ROSIGLITAZONEMALEATE" ~ "Rosiglitazone",
    chemical == "KETOCONAZOLE" ~ "Ketoconazole",
    chemical == "PHENYLBUTAZONE" ~ "Phenylbutazone",
    chemical == "NAPHTHYLISOTHIOCYANATE" ~ "1-Naphthylisothiocyanate",
    chemical == "INTERFERONALPHAHUMAN" ~ "Interferon alpha-2",
    chemical == "PROPYLTHIOURACIL" ~ "Propylthiouracil",
    chemical == "MOXISYLYTE" ~ "Moxisylyte",
    chemical == "DICLOFENAC" ~ "Diclofenac",
    chemical == "NMETHYLNNITROSOUREA" ~ "Methylnitrosourea",
    chemical == "RIFAMPICIN" ~ "Rifampin",
    chemical == "CARBAMAZEPINE" ~ "Carbamazepine",
    chemical == "DIETHYLMALEATE" ~ "diethyl maleate",
    chemical == "PHENOBARBITAL" ~ "Phenobarbital",
    chemical == "FLUPHENAZINE" ~ "Fluphenazine",
    chemical == "PAPAVERINE" ~ "Papaverine",
    chemical == "CAFFEINE" ~ "Caffeine",
    chemical == "FUROSEMIDE" ~ "Furosemide",
    chemical == "NIMESULIDE" ~ "nimesulide",
    chemical == "NITROFURANTOIN" ~ "Nitrofurantoin",
    chemical == "TRANSFORMINGGROWTHFACTORBETA1" ~ "Transforming Growth Factor beta1",
    chemical == "CYCLOPHOSPHAMIDE" ~ "Cyclophosphamide",
    chemical == "LPS" ~ "Lipopolysaccharides",
    chemical == "BUTYLATEDHYDROXYANISOLE" ~ "Butylated Hydroxyanisole",
    chemical == "DILTIAZEM" ~ "Diltiazem",
    chemical == "BROMOETHYLAMINE" ~ "2-bromoethylamine",
    chemical == "TUNICAMYCIN" ~ "Tunicamycin",
    chemical == "TRIAZOLAM" ~ "Triazolam",
    chemical == "SULPIRIDE" ~ "Sulpiride",
    chemical == "SIMVASTATIN" ~ "Simvastatin",
    chemical == "THIORIDAZINE" ~ "Thioridazine",
    TRUE ~ chemical  # fallback to original if no match
  ))
```

```{r}
drw_scores_PHH %>%
  arrange(desc(abs(drw_zscore))) %>%
  distinct(chemical, module_number, .keep_all = TRUE) %>%
  head(100)
```

```{r}
top100_results_PHH_drw <- drw_scores_PHH %>%
  arrange(desc(abs(drw_zscore))) %>%
  distinct(chemical, module_number, .keep_all = TRUE) %>%
  slice_head(n = 100) %>%
  transmute(chemical_ctd,
            module_nr = module_number,
            drw_zscore) %>%    # joining problems
  rowwise() %>%
  mutate(
    module_terms = list(
      module_enrichment_PHH_reactome %>%
        filter(
          module_number == module_nr,       
          source        == "CPDB:Reactome", # to be sure
          padj          <  0.05
        ) %>%
        pull(term) %>%
        unique()
    ),
    ctd_terms = list(
      CTD_Pathways_reactome %>%
        filter(
          `# ChemicalName` == chemical_ctd,
          CorrectedPValue  <  0.05
        ) %>%
        pull(PathwayName) %>%
        unique()
    ),
    overlaps  = list(intersect(module_terms, ctd_terms)),
    n_matches = length(overlaps),
    has_match = as.integer(n_matches > 0)
  ) %>%
  ungroup() %>%
  select(chemical_ctd,
         module_number = module_nr,   
         n_matches,
         has_match,
         drw_zscore)
```


```{r}
print(top100_results_PHH_drw)
```

# RPTEC Analysis

## eg scores

```{r}
eg_scores_RPTEC %>%
  arrange(desc(abs_eg_score)) %>%   
  distinct(chemical, module_number, .keep_all = TRUE) %>%  # top 100 unique combinations
  head(100)
```

```{r}
eg_scores_RPTEC <- eg_scores_RPTEC %>%
  mutate(chemical_ctd = case_when(
    chemical == "DOXORUBICIN" ~ "Doxorubicin",
    chemical == "ARISTOLOCHICACID" ~ "Aristolochic Acids",
    chemical == "LEADACETATE" ~ "lead acetate",
    chemical == "OCHRATOXINA" ~ "ochratoxin A",
    chemical == "MITOMYCINC" ~ "Mitomycin",
    chemical == "CYCLOSPORINA" ~ "Cyclosporins",
    chemical == "RAPAMYCIN" ~ "Sirolimus",
    chemical == "CISPLATIN" ~ "Cisplatin",
    chemical == "ZOLENDRONICACID" ~ "Zoledronic Acid",
    chemical == "INTERLEUKIN1BETA" ~ "Interleukins",
    chemical == "AMPHOTERICINB" ~ "Amphotericin B",
    chemical == "OMEPRAZOL" ~ "Omeprazole",
    TRUE ~ chemical  # fallback to original if no match
  ))
```

```{r}
eg_scores_RPTEC %>%
  arrange(desc(abs_eg_score)) %>%   
  distinct(chemical, module_number, .keep_all = TRUE) %>%  # top 100 unique combinations
  head(100)
```

```{r}
top100_results_RPTEC_eg <- eg_scores_RPTEC %>%
  arrange(desc(abs_eg_score)) %>%
  distinct(chemical_ctd, module_number, .keep_all = TRUE) %>% # Top unique combinations
  slice_head(n = 100) %>%
  transmute(chemical_ctd,
            module_nr = module_number,
            abs_eg_score) %>%    # joining problems
  rowwise() %>%
  mutate(
    module_terms = list(
      module_enrichment_RPTEC_reactome %>%
        filter(
          module_number == module_nr,        # to be sure
          source        == "Reactome_2022",
          padj          <  0.05
        ) %>%
        pull(term_id) %>%
        unique()
    ),
    ctd_terms = list(
      CTD_Pathways_reactome %>%
        filter(
          `# ChemicalName` == chemical_ctd,
          CorrectedPValue  <  0.05
        ) %>%
        pull(reactome_id) %>%
        unique()
    ),
    overlaps  = list(intersect(module_terms, ctd_terms)),
    n_matches = length(overlaps),
    has_match = as.integer(n_matches > 0)
  ) %>%
  ungroup() %>%
  select(chemical_ctd,
         module_number = module_nr,
         n_matches,
         has_match,
         abs_eg_score)
```


```{r}
print(top100_results_RPTEC_eg)
```


## DRW scores

```{r}
drw_scores_RPTEC <- drw_scores_RPTEC %>% 
  filter(chemical != "EPIDERMALGROWTHFACTOR") # No EGF in Database, however, only one entry in top k modules
```

```{r}
drw_scores_RPTEC %>%
  arrange(desc(abs(drw_zscore))) %>%
  distinct(chemical, module_number, .keep_all = TRUE) %>% 
  head(100)
```


```{r}
drw_scores_RPTEC <- drw_scores_RPTEC %>%
  mutate(chemical_ctd = case_when(
    chemical == "INTERLEUKIN1BETA" ~ "Interleukins",
    chemical == "LEADACETATE" ~ "lead acetate",
    chemical == "MITOMYCINC" ~ "Mitomycin",
    chemical == "ZOLENDRONICACID" ~ "Zoledronic Acid",
    chemical == "RAPAMYCIN" ~ "Sirolimus",
    chemical == "AMPHOTERICINB" ~ "Amphotericin B",
    chemical == "ARISTOLOCHICACID" ~ "Aristolochic Acids",
    chemical == "OCHRATOXINA" ~ "ochratoxin A",
    chemical == "CYCLOSPORINA" ~ "Cyclosporins",
    chemical == "DOXORUBICIN" ~ "Doxorubicin",
    chemical == "CISPLATIN" ~ "Cisplatin",
    chemical == "OMEPRAZOL"     ~ "Omeprazole",
    chemical == "TUNICAMYCIN" ~ "Tunicamycin",
    chemical == "PUROMYCIN" ~ "Puromycin",
    chemical == "GENTAMYCIN" ~ "Gentamicins",
    chemical == "PARACETAMOL" ~ "Acetaminophen",
    chemical == "TUMORNECROSEFACTORALFA" ~ "Tumor Necrosis Factors",
    chemical == "CITRININ" ~ "Citrinin",
    TRUE ~ chemical  # fallback to original if no match
  ))
```

```{r}
drw_scores_RPTEC %>%
  arrange(desc(abs(drw_zscore))) %>%
  distinct(chemical, module_number, .keep_all = TRUE) %>%
  head(100)
```

```{r}
top100_results_RPTEC_drw <- drw_scores_RPTEC %>%
  arrange(desc(abs(drw_zscore))) %>%
  distinct(chemical, module_number, .keep_all = TRUE) %>%
  slice_head(n = 100) %>%
  transmute(chemical_ctd,
            module_nr = module_number,
            drw_zscore) %>%    # rename module_nr: problems with joining
  rowwise() %>%
  mutate(
    module_terms = list(
      module_enrichment_RPTEC_reactome %>%
        filter(
          module_number == module_nr,     
          source        == "Reactome_2022", # to be sure
          padj          <  0.05
        ) %>%
        pull(term_id) %>%
        unique()
    ),
    ctd_terms = list(
      CTD_Pathways_reactome %>%
        filter(
          `# ChemicalName` == chemical_ctd,
          CorrectedPValue  <  0.05
        ) %>%
        pull(reactome_id) %>%
        unique()
    ),
    overlaps  = list(intersect(module_terms, ctd_terms)),
    n_matches = length(overlaps),
    has_match = as.integer(n_matches > 0)
  ) %>%
  ungroup() %>%
  select(chemical_ctd,
         module_number = module_nr,
         n_matches,
         has_match,
         drw_zscore)
```


```{r}
print(top100_results_RPTEC_drw)
```

# Final Results

The accuracy for the top k-values will be calculated.

```{r}
# List of top-k values
k_values <- c(20, 40, 60, 80, 100) # for now k is in 20, 40, 60, 80, 100
```


```{r}
# combined results
result_dfs <- list(
  RPTEC_wDRW = top100_results_RPTEC_drw,
  RPTEC_EG  = top100_results_RPTEC_eg,
  PHH_wDRW   = top100_results_PHH_drw,
  PHH_EG    = top100_results_PHH_eg
)
```


```{r}
# Compute accuracy per dataset and k
accuracy_results <- map_dfr(result_dfs, function(df) {
  map_dfr(k_values, function(k) {
    df_k <- df %>% slice_head(n = k)
    tibble(
      k         = k,
      n_matches = sum(df_k$has_match, na.rm = TRUE),
      accuracy  = n_matches / k
    )
  })
}, .id = "dataset")
```

```{r}
accuracy_results <- accuracy_results %>%
  mutate(
    celltype  = ifelse(grepl("PHH", dataset), "PHH", "RPTEC"),
    algorithm = ifelse(grepl("EG", dataset), "Eigengene", "wDRW")
  )
```


```{r}
print(accuracy_results)
```


```{r}
ggplot(accuracy_results, aes(x = algorithm, y = accuracy, fill = factor(k))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Accuracy of Top-k Scoring Modules per Scoring Method",
    x = "Scoring Method",
    y = "Accuracy",
    fill = "Top-k Modules"
  ) +
  facet_grid(. ~ celltype) +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold", size =14)
  )
```
