---
title: "Correlation EG vs wDRW"
author: "Sem"
date: "2025-06-06"
output:
  html_document:
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
      position: left
    toc_depth: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(tibble)
library(dplyr)
library(patchwork)
library(dplyr)
library(tidyverse)
```


# Introduction

In this .Rmd file, the least and most correlated modules between EG and wDRW in PHH and RPTEC are calculated and plotted using dotplots to visualize both module_size and annotation

## PHH data

```{r}
data_dir <- "C:/Users/semde/Documents/BOO_Scripts/Data/PHH_TXG-MAPr"
module_enrichment_PHH   <- readRDS(file.path(data_dir, "module_enrichment.rds"))
drw_scores_PHH <- readRDS(file.path(data_dir, "drw_scores_PHH_r0.7.rds"))
eg_scores_PHH <- readRDS(file.path(data_dir, "eg_score.rds"))
module_annotation_PHH <- readRDS(file.path(data_dir, "module_annotation.rds"))
expression_long_PHH <- readRDS(file.path(data_dir, "expression_long.rds"))
module_definition_PHH <- readRDS(file.path(data_dir, "module_definition.rds"))
GRR_vs_eg_score_PHH <- readRDS(file.path(data_dir, "GRR_vs_eg_score_PHH.rds")) 
```

## RPTEC data

```{r}
data_dir <- "C:/Users/semde/Documents/BOO_Scripts/Data/RPTEC_TXG-MAPr"
module_enrichment_RPTEC   <- readRDS(file.path(data_dir, "module_enrichment.rds"))
drw_scores_RPTEC <- readRDS(file.path(data_dir, "drw_scores_RPTEC_r0.7.rds"))
eg_scores_RPTEC <- readRDS(file.path(data_dir, "eg_score.rds"))
module_annotation_RPTEC <- readRDS(file.path(data_dir, "module_annotation.rds"))
expression_long_RPTEC <- readRDS(file.path(data_dir, "expression_long.rds"))
module_definition_RPTEC <- readRDS(file.path(data_dir, "module_definition.rds"))
GRR_vs_DRW_RPTEC <- readRDS(file.path(data_dir, "GRR_vs_DRW_cor_eg_RPTEC.rds")) 
```

# PHH EG vs wDRW correlation

## Joining

```{r}
PHH_eg_wDRW_df <- eg_scores_PHH %>%
  select(sample_id, module_number, eg_score) %>%
  inner_join(
    drw_scores_PHH %>%
      select(condition_id, module_nr, drw_zscore, module_size),
    by = c("sample_id" = "condition_id", "module_number" = "module_nr")
  )
```

## Plotting

```{r}
ggplot(PHH_eg_wDRW_df, aes(x = eg_score, y = drw_zscore)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  theme_minimal() +
  labs(
    title = "PHH: EG score vs wDRW z-score",
    x = "Eigengene Score (EG)",
    y = "Weighted Directed Random Walk Score (wDRW)"
  )
```

### Calculating Pearson Correlation

```{r}
cor(PHH_eg_wDRW_df$eg_score, PHH_eg_wDRW_df$drw_zscore, method = "pearson")
```

```{r}
module_correlations <- PHH_eg_wDRW_df %>%
  group_by(module_number) %>%
  summarise(
    correlation = cor(eg_score, drw_zscore, method = "pearson"),
    module_size = first(module_size),
    .groups = "drop"
  ) %>%
  mutate(abs_correlation = abs(correlation)) %>% 
  arrange(desc(abs_correlation)) %>% 
  drop_na()
```

```{r}
module_correlations_annotated <- module_correlations %>%
  left_join(
    module_annotation_PHH %>%
      mutate(module_number = as.integer(stringr::str_remove(module, "hPHH_"))) %>%
      select(-module),
    by = "module_number"
  )
```


### Most and least correlated modules

#### Most correlated modules

```{r}
module_correlations_annotated %>% 
  head(10) %>% 
  select(-abs_correlation)
```

#### Least correlated modules

```{r}
module_correlations_annotated %>%
  arrange(abs_correlation) %>% 
  select(-abs_correlation) %>% 
  head(10)
```

### Dotplot

#### Preparing

Annotated modules were summarized for annotation in the plot

```{r}
top10_df <- tribble(
  ~module_number, ~abs_correlation, ~module_size, ~annotation,
  12, 0.9483265, 133, "Immune Response",
  22, 0.9478952, 46, "Immune Response",
  44, 0.9441389, 24, "Immune Response",
  131, 0.9351365, 11, "Heat Shock Response",
  11, 0.9333265, 176, "Metabolism",
  29, 0.9222987, 33, "Extracellular Matrix",
  15, 0.9137702, 91, "Endoplasmic Reticulum",
  37, 0.9128095, 27, "DNA processing",
  134, 0.9081236, 11, "Metabolism",
  358, 0.9014971, 5, "Metabolism"
)
```


```{r}
bottom10_df <- tribble(
  ~module_number, ~abs_correlation, ~module_size, ~annotation,
  246, 0.00993627, 8, NA,
  263, 0.03580652, 7, NA,
  200, 0.05761286, 9, NA,
  348, 0.06589897, 5, NA,
  121, 0.08444078, 12, NA,
  257, 0.09893226, 7, NA,
  207, 0.10470367, 9, NA,
  252, 0.10563228, 7, NA,
  96, 0.11049588, 13, NA,
  32, 0.11170605, 30, NA
)
```

```{r}
combined_df <- bind_rows(top10_df, bottom10_df) %>%
  mutate(group = if_else(module_number %in% top10_df$module_number, "Top correlated", "Least correlated"))
```


#### Plotting

```{r}
annotation_colors <- c(
  "DNA Processing" = "#E674AD",
  "Extracellular Matrix" = "#4682B4",         
  "Immune Response" = "#A0522D",
  "Endoplasmic Reticulum" = "#DD3226",
  "Metabolism" = "#009E73",
  "Heat Shock Response" = "#FF6347",
  "NA" = "#999999"
)
```

```{r}
sorted_modules <- combined_df %>%
  arrange(abs_correlation) %>%
  pull(module_number)

# Zet module_number als factor met gewenste volgorde
combined_df <- combined_df %>%
  mutate(module_number = factor(module_number, levels = sorted_modules))
```


```{r}
# Plot
p <- ggplot(combined_df, aes(x = factor(module_number), 
                             y = abs_correlation, 
                             size = module_size, 
                             color = annotation)) +
  geom_point(alpha = 0.9) +
  scale_color_manual(values = annotation_colors,
                     guide = guide_legend(override.aes = list(size = 10))) +
  scale_size(range = c(10, 20),
             breaks = c(10, 50, 150, 300)) +
  scale_y_continuous(limits = c(0, 1)) +   # Y-as 0 tot 1
  facet_wrap(~group, scales = "free_x", ncol = 2) +  # FACET naast elkaar
  labs(x = "Module Number", 
       y = "Absolute Pearson Correlation", 
       size = "Module Size (n genes)", 
       color = "Annotation") +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(size = 16, face = "bold"),  # facet titles
    axis.text.x = element_text(size = 15, angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15, face = "bold"),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 14),  # annotation labels
    panel.grid.major.x = element_blank(),  # optioneel → schoner
    panel.grid.minor.x = element_blank()
  )

p
```


```{r}
# Save as PDF
ggsave(filename = "EG_wDRW_correlation_PHH.pdf", plot = p, width = 15, height = 8)
```

# RPTEC EG vs wDRW correlation

## Joining

```{r}
RPTEC_eg_wDRW_df <- eg_scores_RPTEC %>%
  select(sample_id, module_number, eg_score) %>%
  inner_join(
    drw_scores_RPTEC %>%
      select(condition_id, module_nr, drw_zscore, module_size),
    by = c("sample_id" = "condition_id", "module_number" = "module_nr")
  )
```

## Plotting

```{r}
ggplot(RPTEC_eg_wDRW_df, aes(x = drw_zscore, y = eg_score)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  theme_minimal() +
  labs(
    title = "RPTEC: EG score vs wDRW z-score",
    x = "Eigengene Score (EG)",
    y = "Weighted Directed Random Walk Score (wDRW)"
  )
```

### Calculating Pearson Correlation

```{r}
cor(RPTEC_eg_wDRW_df$eg_score, RPTEC_eg_wDRW_df$drw_zscore, method = "pearson")
```
```{r}
module_correlations <- RPTEC_eg_wDRW_df %>%
  group_by(module_number) %>%
  summarise(
    correlation = cor(eg_score, drw_zscore, method = "pearson"),
    module_size = first(module_size),
    .groups = "drop"
  ) %>%
  mutate(abs_correlation = abs(correlation)) %>% 
  arrange(desc(abs_correlation)) %>% 
  drop_na()
```

```{r}
module_correlations_annotated <- module_correlations %>%
  left_join(
    module_annotation_RPTEC %>%
      mutate(module_number = as.integer(stringr::str_remove(module, "hRPTECTERT1_"))) %>%
      select(-module),
    by = "module_number"
  )
```


### Most and least correlated modules

#### Most correlated modules

```{r}
# Bekijk result
module_correlations_annotated %>% 
  head(10) %>% 
  select(-abs_correlation)
```

#### Least correlated modules

```{r}
module_correlations_annotated %>%
  arrange(abs_correlation) %>% 
  select(-abs_correlation) %>% 
  head(10)
```

### Dotplot

#### Preparing

```{r}
top10_df <- tribble(
  ~module_number, ~abs_correlation, ~module_size, ~annotation,
  32, 0.9735108, 54, "DNA Processing",
  17, 0.9731647, 129, "Cytoskeleton",
  15, 0.9702463, 142, "Immune Response",
  50, 0.9668785, 24, "Oxidative Stress",
  12, 0.9644864, 162, "Cell Cycle",
  22, 0.9637856, 99, "DNA Processing",
  61, 0.9616752, 20, "Immune Response",
  29, 0.9588111, 73, "Energy",
  33, 0.9581879, 46, "Protein Processing",
  8, 0.9547510, 380, "Energy"
)
```


```{r}
bottom10_df <- tribble(
  ~module_number, ~abs_correlation, ~module_size, ~annotation,
  136, 0.03660615, 10, "Cell Cycle",
  192, 0.05583063, 8, "NA",
  178, 0.15267001, 8, "Extracellular Matrix",
  270, 0.21566644, 6, "Lysosome",
  231, 0.24992180, 7, "NA",
  230, 0.30322487, 7, "Signaling",
  85, 0.33395246, 14, "Oxidative Stress",
  256, 0.34463024, 6, "NA",
  28, 0.35297726, 73, "Protein Processing",
  67, 0.35778764, 17, "NA"
)
```

```{r}
combined_df <- bind_rows(top10_df, bottom10_df) %>%
  mutate(group = if_else(module_number %in% top10_df$module_number, "Top Correlated", "Least Correlated"))
```


#### Plotting

```{r}
annotation_colors <- c(
  "DNA Processing" = "#E674AD",
  "Cytoskeleton" = "#9932CC",         
  "Immune Response" = "#A0522D",
  "Oxidative Stress" = "#3CB44B",
  "Cell Cycle" = "#FFDE17",
  "Mitochondria" = "#39C6F4",
  "Protein Processing" = "#8DA0CB",
  "Extracellular Matrix" = "#4682B4",
  "Lysosome" = "#f58231",
  "Signalling" = "#6a3d9a",     
  "NA" = "#999999"
)
```

```{r}
sorted_modules <- combined_df %>%
  arrange(abs_correlation) %>%
  pull(module_number)

# Zet module_number als factor met gewenste volgorde
combined_df <- combined_df %>%
  mutate(module_number = factor(module_number, levels = sorted_modules))
```


```{r}
# Plot
p <- ggplot(combined_df, aes(x = factor(module_number), 
                             y = abs_correlation, 
                             size = module_size,    
                             color = annotation)) + 
  geom_point(alpha = 0.9) +
  scale_color_manual(values = annotation_colors) +  
  scale_size(range = c(10, 20),
             breaks = c(10, 50, 150, 300)) +
  scale_y_continuous(limits = c(0, 1)) +   
  facet_wrap(~group, scales = "free_x", ncol = 2) +  
  labs(x = "Module Number", 
       y = "Absolute Pearson Correlation", 
       size = "Module Size (n genes)", 
       color = "Annotation") +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 15, angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15, face = "bold"),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 14),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  guides(
    size = guide_legend(order = 1),
    color = guide_legend(order = 2, override.aes = list(size = 10))
  )

p
```


```{r}
# Save as PDF
ggsave(filename = "EG_wDRW_correlation_RPTEC.pdf", plot = p, width = 15, height = 8)
```

## Plotting module 136 (very low cor) and module 32 (very high cor)

## Module 136

### EG vs wDRW in module 136

```{r}
RPTEC_eg_wDRW_df %>%
  filter(module_number == 136) %>%
  ggplot(aes(x = eg_score, y = drw_zscore)) +
  geom_point(size = 3, color = "#1f78b4", alpha = 0.8) +
  geom_smooth(method = "lm", color = "red", se = FALSE, linewidth = 1.2) +
  labs(
    title = "EG score vs wDRW  for Module 136 (very low correlation)",
    x = "Eigengene Score (EG)",
    y = "Weighted Directed Random Walk Score (wDRW)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )
```
### median log2fc vs wDRW and EG in module 136

```{r}
grr_df <- GRR_vs_DRW_RPTEC %>%
  filter(module_number == 136) %>%
  select(sample_id, module_number, wGRR, GRR)
```


```{r}
scores_df <- RPTEC_eg_wDRW_df %>%
  filter(module_number == 136) %>%
  select(sample_id, module_number, eg_score, drw_zscore)
```


```{r}
module136_combined <- scores_df %>%
  left_join(grr_df, by = c("sample_id", "module_number"))
```


```{r}
module136_long <- module136_combined %>%
  pivot_longer(cols = c("eg_score", "drw_zscore"), 
               names_to = "score_type", 
               values_to = "score")
```


```{r}
p1 <- ggplot(module136_combined, aes(x = eg_score, y = GRR)) +
  geom_point(alpha = 0.8, size = 3, color = "#1f78b4") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(title = "EG and wDRW score vs GRR (Module 136)",
       x = "EG_score",
       y = "GRR") +
  theme_minimal(base_size = 14)

# Plot DRW
p2 <- ggplot(module136_combined, aes(x = drw_zscore, y = GRR)) +
  geom_point(alpha = 0.8, size = 3, color = "#33a02c") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(title = "",
       x = "wDRW",
       y = "GRR") +
  theme_minimal(base_size = 14)

p1 + p2 + plot_layout(ncol = 2)
```

```{r}
p1 <- ggplot(module136_combined, aes(x = eg_score, y = wGRR)) +
  geom_point(alpha = 0.8, size = 3, color = "red") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(title = "EG and wDRW score vs wGRR (Module 136)",
       x = "EG_score",
       y = "GRR") +
  theme_minimal(base_size = 14)

# Plot DRW
p2 <- ggplot(module136_combined, aes(x = drw_zscore, y = wGRR)) +
  geom_point(alpha = 0.8, size = 3, color = "purple") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(title = "",
       x = "wDRW",
       y = "wGRR") +
  theme_minimal(base_size = 14)

p1 + p2 + plot_layout(ncol = 2)
```

## Module 32

```{r}
RPTEC_eg_wDRW_df %>%
  filter(module_number == 32) %>%
  ggplot(aes(x = eg_score, y = drw_zscore)) +
  geom_point(size = 3, color = "#1f78b4", alpha = 0.8) +
  geom_smooth(method = "lm", color = "red", se = FALSE, linewidth = 1.2) +
  labs(
    title = "EG score vs wDRW  for Module 32 (very high correlation)",
    x = "Eigengene Score (EG)",
    y = "Weighted Directed Random Walk Score (wDRW)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )
```



