---
title: "PHH_EG"
author: "Sem"
date: "2025-05-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
```

```{r}
data_dir <- "C:/Users/semde/Documents/BOO_Scripts/Data/PHH_TXG-MAPr"
expression_long   <- readRDS(file.path(data_dir, "expression_long.rds"))
module_definition        <- readRDS(file.path(data_dir, "module_definition.rds"))
eg_score          <- readRDS(file.path(data_dir, "eg_score.rds"))
module_annotation <- readRDS(file.path(data_dir, "module_annotation.rds"))
```

# Introduction

The eigengene scoring method will be evaluated in the Primary Human Hepatocyte (PHH) dataset from the TXG-MAPr tool. It will be evaluated on the weighted Gene Response Rate (wGRR) metric, which is proposed as an evaluation metric for the topological activation of every unique combination of condition and module. The wGRR is presented as the following formula:


$$
GRR(M_j) = 
\frac{
  \displaystyle \sum_{i = 1}^{n_j}
  \left( 
    \left| \log_2 \mathrm{FC}_i \right| > 1.0 \land p_{\mathrm{adj},i} < 0.05 
  \right)
  \cdot 
  \mathrm{sign} \left( \log_2 \mathrm{FC}_i \right)
}{
  \displaystyle n_j
}
$$


$$
wGRR(M_j) = \sum_{i = 1}^{n_j}
\left( 
  \left| \log_2 \mathrm{FC}_i \right| > 1.0 \land p_{\mathrm{adj},i} < 0.05 
\right)
\cdot 
\frac{ \mathrm{median(TOM)}_i }{ \sum_{i=1}^{n_j} \mathrm{median(TOM)} }
\cdot 
\mathrm{sign} \left( \log_2 \mathrm{FC}_i \right)
$$
Here, the binary outcome (0 or 1) for the significance filter is weighted by the cor_eg of each gene. And the total weighted significant genes are divided by the total genes in the module.

# Distribution of Log2FC and adjusted p-value

Distributions will be visualized for insights on the data

## Log2FC

```{r}
ggplot(expression_long, aes(x = log2fc)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "steelblue", alpha = 0.6) +
  geom_density(color = "darkblue") +
  theme_minimal() + labs(title = "Distribution of Log2FC in PHH dataset")
```

```{r}
sum(abs(expression_long$log2fc) > 1)
```


## p-adjusted

```{r}
ggplot(expression_long, aes(x = padj)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "darkgreen", alpha = 0.6) +
  geom_density(color = "black") +
  theme_minimal() +
  labs(title = "Distribution of adjusted p-values in PHH dataset")
```

```{r}
sum(expression_long$padj <  0.05)
```

## Both Log2FC and padj significant

How many significant genes in the PHH dataset

```{r}
sum(abs(expression_long$log2fc) > 1 & expression_long$padj < 0.05)
```

# Gene Response Rate (GRR)

**This section is not relevant anymore, because now the wGRR is used and not the GRR. However, it is still included in the .Rmd file, if it is decided to use the GRR metric in the future**

The proposed Gene Response Rate reflects the total of significant genes (abs(Log2FC) > 1.0 & padj < 0.05) in a module with size n. This will be plotted against the eigengene score to assess this scoring method.

$$
GRR = \frac{ \displaystyle \sum_{i=1}^{n} \left( |\log_2 \mathrm{FC}_i| > 1.0 \ \land \ p_{\mathrm{adj},i} < 0.05 \right) }{ n }
$$

## Merge module_definition with expression_long

Add module_definition to the genes

```{r}
# Add the modules to the genes in expression_long, don't join with all the variabels -> saves memory

expression_with_module <- expression_long %>%
  select(entrez_id, sample_id, log2fc, padj) %>%
  inner_join(
    module_definition %>% select(entrez_id, module_number, module_size),
    by = "entrez_id"
  ) %>%
  filter(module_number != 0) # Module 0 is irrelevant so delete
```

## Calculate significant genes and total per module

Determine what genes are significant (log2fc > 1.0 and padj < 0.05)

```{r}
# Calculate significant genes
expression_with_module <- expression_with_module %>%
  mutate(significant = abs(log2fc) > 1 & padj < 0.05)


# Count significant genes per module
sig_count <- expression_with_module %>%
  group_by(module_number, sample_id) %>%
  summarise(sig_genes = sum(significant), .groups = "drop")
```

## Add the module_size to sig_count and calculate GRR

Divide the total significant genes by the module size

```{r}
sig_count <- sig_count %>%
  left_join(
    expression_with_module %>%
      distinct(module_number, module_size),
    by = "module_number"
  )
```

```{r}
sig_count <- sig_count %>%
  mutate(GRR = sig_genes / module_size)
```

## Join GRR with EG score per sample 

Make a df with the EG_score and corresponding GRR

```{r}
GRR_vs_eg_score <- sig_count %>%
  inner_join(eg_score, by = c("module_number", "sample_id"))
```

# Weighted Gene Response Rate (wGRR)

Here the wGRR is calculated for each combination of condition and module in the PHH dataset.

```{r}
weighted_grr_df <- expression_with_module %>%
  # remove module 0
  filter(module_number != 0) %>%
  
  # double module_size
  select(-module_size) %>%
  
  # Add per gen corr_egs + module_size
  left_join(
    module_definition %>% select(entrez_id, module_number, corr_egs, module_size),
    by = c("entrez_id", "module_number")
  ) %>%
  
  # Calculate if the gene is significantly expressed
  mutate(is_significant = as.numeric(abs(log2fc) > 1 & padj < 0.05)) %>%
  
  # calculate wGRR per combiantion of sample and module
  group_by(sample_id, module_number) %>%
  summarise(
    wGRR = sum(is_significant * abs(corr_egs), na.rm = TRUE),
    module_size = first(module_size),
    .groups = "drop"
  ) %>%
  
  # Normalize for module_size
  mutate(wGRR = wGRR / module_size)
```


```{r}
# Join the calculated wGRR with GRR_vs_eg_score
GRR_vs_eg_score <- GRR_vs_eg_score %>%
  select(-module_size) %>%
  left_join(weighted_grr_df, by = c("sample_id", "module_number"))
```

```{r}
sum(is.na(GRR_vs_eg_score$wGRR))
```

```{r}
min(GRR_vs_eg_score$wGRR)
```

```{r}
#saveRDS(GRR_vs_eg_score, file = "GRR_vs_eg_score_PHH.rds")
```

## Density plots of GRR and EG score

Inspect the distribution of the GRR, EG and abs EG.

```{r}
ggplot(GRR_vs_eg_score, aes(x = wGRR)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "green", alpha = 0.6) +
  geom_density(color = "black") +
  theme_minimal() + labs(title = "Distribution of Weighted Gene Response Rate in PHH dataset")
```

Like expected, the wGRR shows a logaritmic distribution due to the count data nature.

```{r}
ggplot(GRR_vs_eg_score, aes(x = eg_score)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "red", alpha = 0.6) +
  geom_density(color = "black") +
  theme_minimal() + labs(title = "Distribution of Eigengene Score in PHH dataset")
```

Eigengene shows a normal distribution, as expected due to the PCA use.

```{r}
ggplot(GRR_vs_eg_score, aes(x = abs_eg_score)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "red", alpha = 0.6) +
  geom_density(color = "black") +
  theme_minimal() + labs(title = "Distribution of Absolute Eigengene Score in PHH dataset")
```

Abs Eigengene shows a logarthmic distribution

## Visualize GRR vs EG score with a linear fit

```{r}
ggplot(GRR_vs_eg_score, aes(x = abs_eg_score, y = wGRR)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  theme_minimal() +
  labs(
    title = " Weighted Gene Response Rate vs Abs Eigengene Score in PHH",
    x = "Absolute Eigengene Score (EG)",
    y = "Weighted Gene Response Rate (wGRR)"
  )
```

The linear fit does not follow the trend of the wGRR vs EG, as expected due to the logarithmic distribution of the wGRR. Next we will try to fit a different, logarthmic fit. To detect significant outliers.

## Binomial Generalized Linear Model (GLM)

The data will be fitted with a binomial GLM to assess the general trend and later determine outlying modules from this trend


### Fitting

```{r}
glm_data <- GRR_vs_eg_score %>%
  filter(!is.na(wGRR), !is.na(abs_eg_score)) # This filtering was performed as the fit was first tested on a subset, but the filtering is irrelevant for the whole dataset as there are no NAs
```


```{r}
glm_fit <- glm(
  wGRR ~ abs_eg_score,
  data = glm_data,
  family = quasibinomial(link = "logit") # quasibinomial as the wGRR is a proportion multiplied by a continuous variable, the cor_eg.
)
```

```{r}
# Make predictions and residuals for determining outliers later
glm_predictions <- predict(glm_fit, type = "response")
glm_dev_resid <- residuals(glm_fit, type = "deviance")

# Combine into a new dataframe
GRR_glm_fit_df <- GRR_vs_eg_score %>%
  mutate(
    fitted_prob = glm_predictions,
    fitted_sig_genes = fitted_prob * module_size,
    residual = sig_genes - fitted_sig_genes,
    dev_resid = glm_dev_resid
  )
```

### Evaluating the fit

```{r}
ggplot(GRR_glm_fit_df, aes(x = abs_eg_score, y = wGRR)) +
  geom_point(alpha = 0.5, size = 0.8) +
  geom_line(aes(y = fitted_sig_genes / module_size), color = "blue", size = 1) +
  theme_minimal() +
  labs(
    title = "Binomial GLM fit: wGRR vs Abs Eigengene Score in PHH",
    x = "Absolute Eigengene Score",
    y = "Weighted Gene Response Rate (wGRR)"
  )
```

This trend seems to fit the data better as expected

```{r}
summary(glm_fit)
```

```{r}
deviance(glm_fit) / df.residual(glm_fit)
```

```{r}
1 - (glm_fit$deviance / glm_fit$null.deviance)
```

### Significant outliers 

For each module and sample, we compare the observed number of significant genes to the expected number based on the binomial GLM, and calculate a two-sided z-score and p-value using the binomial standard error. By doing this, the significant outlying modules from the glm fit are identified. Above and Below fit genes are identified for later analysis.

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(
    se_fit = sqrt(module_size * fitted_prob * (1 - fitted_prob)),
    z_score = (sig_genes - fitted_sig_genes) / se_fit,
    p_value = 2 * pnorm(-abs(z_score))  # two-sided test
  )
```

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(outlier = p_value < 0.05)
```

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(
    outlier_direction = case_when(
      outlier & z_score > 0 ~ "Above fit",
      outlier & z_score < 0 ~ "Below fit",
      TRUE ~ "Not significant"
    )
  )
```

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(module = paste0("hPHH_", module_number)) %>%
  left_join(module_annotation, by = "module")
```

```{r}
sum(GRR_glm_fit_df$outlier, na.rm = TRUE)
```

a total of 11873 outliers from the fit are found.

```{r}
ggplot(GRR_glm_fit_df, aes(x = abs_eg_score, y = wGRR)) +
  geom_point(aes(color = outlier), alpha = 0.6, size = 0.9) +
  geom_line(aes(y = fitted_sig_genes / module_size), color = "blue", size = 1) +
  scale_color_manual(values = c("black", "red")) +
  theme_minimal() +
  labs(
    title = "wGRR vs Abs EG Score with Significant Outliers in PHH",
    x = "Absolute Eigengene Score (EG))",
    y = "Weighted Gene Response Rate (GRR)"
  )
```

This plot shows the fit with the identified outliers in red

## Modules and conditions with most significant outliers'

Based on the binomial GLM fit, we will now find the modules and conditions with the most significant outliers

```{r}
significant_outliers_filtered <- GRR_glm_fit_df %>%
  filter(outlier)
```

```{r}
outliers_per_module <- significant_outliers_filtered %>%
  count(module_number, sort = TRUE)
```

```{r}
outliers_per_sample <- significant_outliers_filtered %>%
  count(sample_id, sort = TRUE)
```

```{r}
head(outliers_per_module, 10) # Top modules with outliers
head(outliers_per_sample, 10) # Top cconditions with outliers
```

### Outliers per module: total, below and above the fit with human annotation

Here, we will visualize the modules with the most outliers. Total, below and above the Binomial GLM fit.

#### Total outliers from the GLM fit

```{r}
threshold <- 300  # Threshold for annotation

GRR_glm_fit_df %>%
  filter(outlier) %>%
  count(module, annotation, sort = TRUE) %>%
  slice_max(n, n = 20, with_ties = FALSE) %>%
  mutate(
    max_n = max(n),
    hjust_val = ifelse(n < threshold, 0, 1.1),
    label_pos = ifelse(n < threshold, n + 2, n - 2)
  ) %>%
  ggplot(aes(x = reorder(module, n))) +
  geom_col(aes(y = max_n), fill = "grey97", width = 0.8) +
  geom_col(aes(y = n), fill = "lightblue", width = 0.8) +
  geom_text(
    aes(y = label_pos, label = annotation, hjust = hjust_val),
    size = 3, color = "black"
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(
    limits = c(0, max(GRR_glm_fit_df %>%
      filter(outlier) %>%
      count(module) %>%
      slice_max(n, n = 20, with_ties = FALSE) %>%
      pull(n)) + 10),
    expand = c(0, 0)
  ) +
  theme_minimal() +
  theme(
    plot.margin = margin(5.5, 100, 5.5, 5.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Top 20 modules with TOTAL significant GRR outliers (PHH)",
    x = "Module",
    y = "Number of significant outliers"
  )
```

#### Outliers above the GLM fit

```{r}
threshold <- 0  # Threshold for switching annotation position

GRR_glm_fit_df %>%
  filter(outlier_direction == "Above fit") %>%
  count(module, annotation, sort = TRUE) %>%
  slice_max(n, n = 20, with_ties = FALSE) %>%
  mutate(
    max_n = max(n),
    hjust_val = ifelse(n < threshold, 0, 1.1),
    label_pos = ifelse(n < threshold, n + 2, n - 2)
  ) %>%
  ggplot(aes(x = reorder(module, n))) +
  geom_col(aes(y = max_n), fill = "grey97", width = 0.8) +
  geom_col(aes(y = n), fill = "lightgreen", width = 0.8) +
  geom_text(
    aes(y = label_pos, label = annotation, hjust = hjust_val),
    size = 3, color = "black"
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(
    limits = c(0, max(GRR_glm_fit_df %>%
      filter(outlier_direction == "Above fit") %>%
      count(module) %>%
      slice_max(n, n = 20, with_ties = FALSE) %>%
      pull(n)) + 10),
    expand = c(0, 0)
  ) +
  theme_minimal() +
  theme(
    plot.margin = margin(5.5, 100, 5.5, 5.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Top 20 modules with outliers ABOVE the GLM fit (PHH)",
    x = "Module",
    y = "Number of outliers above fit"
  )
```

#### Outliers below the GLM fit

```{r}
threshold <- 200  # Threshold for annotation in plot

GRR_glm_fit_df %>%
  filter(outlier_direction == "Below fit") %>%
  count(module, annotation, sort = TRUE) %>%
  slice_max(n, n = 20, with_ties = FALSE) %>%
  mutate(
    max_n = max(n),
    hjust_val = ifelse(n < threshold, 0, 1.1),       # outside or inside
    label_pos = ifelse(n < threshold, n + 2, n - 2)  # offset label position
  ) %>%
  ggplot(aes(x = reorder(module, n))) +
  
  # Background bar
  geom_col(aes(y = max_n), fill = "grey97", width = 0.8) +
  
  # Foreground bar (actual value)
  geom_col(aes(y = n), fill = "pink", width = 0.8) +
  
  # Smart annotation positioning
  geom_text(
    aes(y = label_pos, label = annotation, hjust = hjust_val),
    size = 3, color = "black"
  ) +
  
  coord_flip(clip = "off") +  # Flip axes for horizontal bars

  # Y-axis limits: no exessive white space
  scale_y_continuous(limits = c(0, max(GRR_glm_fit_df %>%
    filter(outlier_direction == "Below fit") %>%
    count(module) %>%
    slice_max(n, n = 20, with_ties = FALSE) %>%
    pull(n)) + 10)) +

  theme_minimal() +
  theme(
    plot.margin = margin(5.5, 100, 5.5, 5.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Top 20 modules with outliers BELOW the GLM fit (PHH)",
    x = "Module",
    y = "Number of outliers below fit"
  )
```

### Outliers per sample: above, below and total.

Now, we will assess what conditions have the most outliers. Total, below and above the binomial GLM fit.

```{r}
outliers_per_sample %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(sample_id, n), y = n)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  theme_minimal() +
  ggtitle("Top 20 conditions with significant GRR outliers from the binomial GLM fit (PHH)") +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0),
    plot.margin = margin(20, 20, 10, 40) # top, right, bottom, left
  ) +
  labs(
    x = "Condition (sample_id)",
    y = "Number of outliers"
  )
```

```{r}
GRR_glm_fit_df %>%
  filter(outlier_direction == "Above fit") %>%
  count(sample_id, sort = TRUE) %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(sample_id, n), y = n)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  theme_minimal() +
  ggtitle("Top 20 conditions with outliers ABOVE the binomial GLM fit (PHH)") +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0),
    plot.margin = margin(20, 20, 10, 40)
  ) +
  labs(
    x = "Condition (sample_id)",
    y = "Number of outliers above fit"
  )
```

```{r}
GRR_glm_fit_df %>%
  filter(outlier_direction == "Below fit") %>%
  count(sample_id, sort = TRUE) %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(sample_id, n), y = n)) +
  geom_col(fill = "purple") +
  coord_flip() +
  theme_minimal() +
  ggtitle("Top 20 conditions with outliers BELOW the binomial GLM fit (PHH)") +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0),
    plot.margin = margin(20, 20, 10, 40)
  ) +
  labs(
    x = "Condition (sample_id)",
    y = "Number of outliers below fit"
  )
```

# Outliers per number of modules * number of samples: density of outliers for comparison with other datasets and techniques.

To compare this with RPTEC the number of outliers will be compared relative to the total modules and conditions. Also, to have a more comparable number to compare with the other topology-based scoring methods

```{r}
total_modules_all <- GRR_glm_fit_df %>%
  filter(module_number != 0) %>%
  summarise(n = n_distinct(module_number)) %>%
  pull(n) # for normalization
```


```{r}
total_conditions_all <- GRR_glm_fit_df %>%
  filter(module_number != 0) %>%
  summarise(n = n_distinct(sample_id)) %>%
  pull(n) # for normalization
```


```{r}
# Making a summarizing df for comparing with other datasets and scoring algorithms
outlier_density <- GRR_glm_fit_df %>%
  filter(outlier, module_number != 0) %>%
  group_by(outlier_direction) %>%
  summarise(
    total_outliers = n(),
    .groups = "drop"
  ) %>%
  bind_rows(
    tibble(
      outlier_direction = "Total",
      total_outliers = sum(.$total_outliers)
    )
  ) %>%
  mutate(
    total_modules = total_modules_all,
    total_conditions = total_conditions_all,
    density = total_outliers / (total_modules * total_conditions)
  )
```

```{r}
print(outlier_density)
```

This shows the total outliers above and below the GLM fit. This is normalized to the total number of combinations of modules and samples to gain the density of the outliers. This is proposed as a quantification for the robustness of EG scoring on the wGRR metric. These results will be shown in a stacked bar plot to compare with the other scoring methods and datasets.
