---
title: "PHH_EG"
author: "Sem"
date: "2025-05-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
```
```{r}
data_dir <- "C:/Users/semde/Universiteit Leiden/BOO 2025 - BOO CDS Giulia team - BOO CDS Giulia team/Students/Sem/Data/PHH_TXG-MAPr"
expression_long   <- readRDS(file.path(data_dir, "expression_long.rds"))
module_definition        <- readRDS(file.path(data_dir, "module_definition.rds"))
eg_score          <- readRDS(file.path(data_dir, "eg_score.rds"))
module_annotation <- readRDS(file.path(data_dir, "module_annotation.rds"))
```



# Distribution of Log2FC and adjusted p-value

Distributions will be visualized for insights on the data

## Log2FC

```{r}
ggplot(expression_long, aes(x = log2fc)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "steelblue", alpha = 0.6) +
  geom_density(color = "darkblue") +
  theme_minimal() + labs(title = "Distribution of Log2FC in PHH dataset")
```

```{r}
sum(abs(expression_long$log2fc) > 1)
```


## p-adjusted

```{r}
ggplot(expression_long, aes(x = padj)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "darkgreen", alpha = 0.6) +
  geom_density(color = "black") +
  theme_minimal() +
  labs(title = "Distribution of adjusted p-values in PHH dataset")
```

```{r}
sum(expression_long$padj <  0.05)
```

## Both Log2FC and padj significant

How many significant genes in the PHH dataset

```{r}
sum(abs(expression_long$log2fc) > 1 & expression_long$padj < 0.05)
```

# Gene Response Rate

The proposed Gene Response Rate reflects the total of significant genes (abs(Log2FC) > 1.0 & padj < 0.05) in a module with size n. This will be plotted against the eigengene score to assess this scoring method.

$$
GRR = \frac{ \displaystyle \sum_{i=1}^{n} \left( |\log_2 \mathrm{FC}_i| > 1.0 \ \land \ p_{\mathrm{adj},i} < 0.05 \right) }{ n }
$$
## Merge module_definition with expression_long

Add module_definition to the genes

```{r}
# Add the modules to the genes in expression_long, don't join with all the variabels -> saves memory

expression_with_module <- expression_long %>%
  select(entrez_id, sample_id, log2fc, padj) %>%
  inner_join(
    module_definition %>% select(entrez_id, module_number, module_size),
    by = "entrez_id"
  ) %>%
  filter(module_number != 0) # Module 0 is irrelevant so delete
```

## Calculate significant genes and total per module

Determine what genes are siginificant (log2fc > 1.0 and padj < 0.05)

```{r}
# Calculate significant genes
expression_with_module <- expression_with_module %>%
  mutate(significant = abs(log2fc) > 1 & padj < 0.05)


# Count significant genes per module
sig_count <- expression_with_module %>%
  group_by(module_number, sample_id) %>%
  summarise(sig_genes = sum(significant), .groups = "drop")
```

## Add the module_size to sig_count and calculate GRR

Divide the total significant genes by the module size

```{r}
sig_count <- sig_count %>%
  left_join(
    expression_with_module %>%
      distinct(module_number, module_size),
    by = "module_number"
  )
```

```{r}
sig_count <- sig_count %>%
  mutate(GRR = sig_genes / module_size)
```

## Join GRR with EG score per sample 

Make a df with the EG_score and corresponding GRR

```{r}
GRR_vs_eg_score <- sig_count %>%
  inner_join(eg_score, by = c("module_number", "sample_id"))
```

## Density plots of GRR and EG score

Inspect the distribution of the GRR, EG and abs EG.

```{r}
ggplot(GRR_vs_eg_score, aes(x = GRR)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "green", alpha = 0.6) +
  geom_density(color = "black") +
  theme_minimal() + labs(title = "Distribution of Gene Response Rate in PHH dataset")
```
Like expected, the GRR shows a logaritmic distribution due to the count data nature.

```{r}
ggplot(GRR_vs_eg_score, aes(x = eg_score)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "red", alpha = 0.6) +
  geom_density(color = "black") +
  theme_minimal() + labs(title = "Distribution of Eigengene Score in PHH dataset")
```
Eigengene shows a normal distribution, as expected due to the PCA use.

```{r}
ggplot(GRR_vs_eg_score, aes(x = abs_eg_score)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "red", alpha = 0.6) +
  geom_density(color = "black") +
  theme_minimal() + labs(title = "Distribution of Absolute Eigengene Score in PHH dataset")
```

## Visualize GRR vs EG score with a linear fit

```{r}
ggplot(GRR_vs_eg_score, aes(x = abs_eg_score, y = GRR)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  theme_minimal() +
  labs(
    title = "Gene Response Rate vs Eigengene Score in PHH",
    x = "Absolute Eigengene Score (EG)",
    y = "Gene Response Rate (GRR)"
  )
```

The linear fit does not follow the trend of the GRR vs EG, as expected due to the logarithmic distribution of the GRR. Next we will try to fit a different, logarthmic fit. To detect significant outliers.

## Binomial Generalized Linear Model (GLM)

The data will be fitted with a binomial GLM to assess the general trend and later determine outlying modules from this trend

```{r}
glm_fit <- glm(
  cbind(sig_genes, module_size - sig_genes) ~ abs_eg_score,
  family = binomial(link = "logit"),
  data = GRR_vs_eg_score
)
```

```{r}
# Make predictions and residuals
glm_predictions <- predict(glm_fit, type = "response")
glm_dev_resid <- residuals(glm_fit, type = "deviance")

# Combine into a new dataframe
GRR_glm_fit_df <- GRR_vs_eg_score %>%
  mutate(
    fitted_prob = glm_predictions,
    fitted_sig_genes = fitted_prob * module_size,
    residual = sig_genes - fitted_sig_genes,
    dev_resid = glm_dev_resid
  )
```

```{r}
ggplot(GRR_glm_fit_df, aes(x = abs_eg_score, y = GRR)) +
  geom_point(alpha = 0.5, size = 0.8) +
  geom_line(aes(y = fitted_sig_genes / module_size), color = "blue", size = 1) +
  theme_minimal() +
  labs(
    title = "Binomial GLM fit: GRR vs Eigengene Score in PHH",
    x = "Eigengene Score",
    y = "Gene Response Rate (GRR)"
  )
```
```{r}
summary(glm_fit)
```

```{r}
deviance(glm_fit) / df.residual(glm_fit)
```

```{r}
1 - (glm_fit$deviance / glm_fit$null.deviance)
```

### significant outliers 

For each module and sample, we compare the observed number of significant genes to the expected number based on the binomial GLM, and calculate a two-sided z-score and p-value using the binomial standard error. By doing this, the significant outlying modules from the glm fit are identified.

Also, it will be determined if the outliers are below or above the fit. And the human annotation will be added.

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(
    se_fit = sqrt(module_size * fitted_prob * (1 - fitted_prob)),
    z_score = (sig_genes - fitted_sig_genes) / se_fit,
    p_value = 2 * pnorm(-abs(z_score))  # two-sided test
  )
```

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(outlier = p_value < 0.05)
```

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(
    outlier_direction = case_when(
      outlier & z_score > 0 ~ "Above fit",
      outlier & z_score < 0 ~ "Below fit",
      TRUE ~ "Not significant"
    )
  )
```

```{r}
GRR_glm_fit_df <- GRR_glm_fit_df %>%
  mutate(module = paste0("hPHH_", module_number)) %>%
  left_join(module_annotation, by = "module")
```


```{r}
sum(GRR_glm_fit_df$outlier, na.rm = TRUE)
```


```{r}
ggplot(GRR_glm_fit_df, aes(x = abs_eg_score, y = GRR)) +
  geom_point(aes(color = outlier), alpha = 0.6, size = 0.9) +
  geom_line(aes(y = fitted_sig_genes / module_size), color = "blue", size = 1) +
  scale_color_manual(values = c("black", "red")) +
  theme_minimal() +
  labs(
    title = "GRR vs Eigengene Score with Significant Outliers in PHH",
    x = "Eigengene Score",
    y = "Gene Response Rate (GRR)"
  )
```

## Modules and conditions with most significant outliers'

Based on the binomial GLM fit, we will now find the modules and conditions with the most significant outliers

```{r}
significant_outliers_filtered <- GRR_glm_fit_df %>%
  filter(outlier)
```

```{r}
outliers_per_module <- significant_outliers_filtered %>%
  count(module_number, sort = TRUE)
```

```{r}
outliers_per_sample <- significant_outliers_filtered %>%
  count(sample_id, sort = TRUE)
```

```{r}
head(outliers_per_module, 10) # Top modules with outliers
head(outliers_per_sample, 10) # Top cconditions with outliers
```
### Outliers per module: total, below and above the fit with human annotation

Here, we will visualize the modules with the most outliers. Total, below and above the Binomial GLM fit.

#### Total outliers from the GLM fit

```{r}
threshold <- 300  # Same threshold

GRR_glm_fit_df %>%
  filter(outlier) %>%
  count(module, annotation, sort = TRUE) %>%
  slice_max(n, n = 20, with_ties = FALSE) %>%
  mutate(
    max_n = max(n),
    hjust_val = ifelse(n < threshold, 0, 1.1),
    label_pos = ifelse(n < threshold, n + 2, n - 2)
  ) %>%
  ggplot(aes(x = reorder(module, n))) +
  geom_col(aes(y = max_n), fill = "grey97", width = 0.8) +
  geom_col(aes(y = n), fill = "lightblue", width = 0.8) +
  geom_text(
    aes(y = label_pos, label = annotation, hjust = hjust_val),
    size = 3, color = "black"
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(
    limits = c(0, max(GRR_glm_fit_df %>%
      filter(outlier) %>%
      count(module) %>%
      slice_max(n, n = 20, with_ties = FALSE) %>%
      pull(n)) + 10),
    expand = c(0, 0)
  ) +
  theme_minimal() +
  theme(
    plot.margin = margin(5.5, 100, 5.5, 5.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Top 20 modules with TOTAL significant GRR outliers (PHH)",
    x = "Module",
    y = "Number of significant outliers"
  )
```

#### Outliers above the GLM fit

```{r}
threshold <- 0  # Threshold for switching annotation position

GRR_glm_fit_df %>%
  filter(outlier_direction == "Above fit") %>%
  count(module, annotation, sort = TRUE) %>%
  slice_max(n, n = 20, with_ties = FALSE) %>%
  mutate(
    max_n = max(n),
    hjust_val = ifelse(n < threshold, 0, 1.1),
    label_pos = ifelse(n < threshold, n + 2, n - 2)
  ) %>%
  ggplot(aes(x = reorder(module, n))) +
  geom_col(aes(y = max_n), fill = "grey97", width = 0.8) +
  geom_col(aes(y = n), fill = "lightgreen", width = 0.8) +
  geom_text(
    aes(y = label_pos, label = annotation, hjust = hjust_val),
    size = 3, color = "black"
  ) +
  coord_flip(clip = "off") +
  scale_y_continuous(
    limits = c(0, max(GRR_glm_fit_df %>%
      filter(outlier_direction == "Above fit") %>%
      count(module) %>%
      slice_max(n, n = 20, with_ties = FALSE) %>%
      pull(n)) + 10),
    expand = c(0, 0)
  ) +
  theme_minimal() +
  theme(
    plot.margin = margin(5.5, 100, 5.5, 5.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Top 20 modules with outliers ABOVE the GLM fit (PHH)",
    x = "Module",
    y = "Number of outliers above fit"
  )
```

#### Outliers below the GLM fit

```{r}
threshold <- 200  # Threshold for annotation in plot

GRR_glm_fit_df %>%
  filter(outlier_direction == "Below fit") %>%
  count(module, annotation, sort = TRUE) %>%
  slice_max(n, n = 20, with_ties = FALSE) %>%
  mutate(
    max_n = max(n),
    hjust_val = ifelse(n < threshold, 0, 1.1),       # outside or inside
    label_pos = ifelse(n < threshold, n + 2, n - 2)  # offset label position
  ) %>%
  ggplot(aes(x = reorder(module, n))) +
  
  # Background bar
  geom_col(aes(y = max_n), fill = "grey97", width = 0.8) +
  
  # Foreground bar (actual value)
  geom_col(aes(y = n), fill = "pink", width = 0.8) +
  
  # Smart annotation positioning
  geom_text(
    aes(y = label_pos, label = annotation, hjust = hjust_val),
    size = 3, color = "black"
  ) +
  
  coord_flip(clip = "off") +  # Flip axes for horizontal bars

  # Y-axis limits: no exessive white space
  scale_y_continuous(limits = c(0, max(GRR_glm_fit_df %>%
    filter(outlier_direction == "Below fit") %>%
    count(module) %>%
    slice_max(n, n = 20, with_ties = FALSE) %>%
    pull(n)) + 10)) +

  theme_minimal() +
  theme(
    plot.margin = margin(5.5, 100, 5.5, 5.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Top 20 modules with outliers BELOW the GLM fit (PHH)",
    x = "Module",
    y = "Number of outliers below fit"
  )
```

### Outliers per sample: above, below and total.

Now, we will assess what conditions have the most outliers. Total, below and above the binomial GLM fit.

```{r}
outliers_per_sample %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(sample_id, n), y = n)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  theme_minimal() +
  ggtitle("Top 20 conditions with significant GRR outliers from the binomial GLM fit (PHH)") +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0),
    plot.margin = margin(20, 20, 10, 40) # top, right, bottom, left
  ) +
  labs(
    x = "Condition (sample_id)",
    y = "Number of outliers"
  )
```

```{r}
GRR_glm_fit_df %>%
  filter(outlier_direction == "Above fit") %>%
  count(sample_id, sort = TRUE) %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(sample_id, n), y = n)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  theme_minimal() +
  ggtitle("Top 20 conditions with outliers ABOVE the binomial GLM fit (PHH)") +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0),
    plot.margin = margin(20, 20, 10, 40)
  ) +
  labs(
    x = "Condition (sample_id)",
    y = "Number of outliers above fit"
  )
```

```{r}
GRR_glm_fit_df %>%
  filter(outlier_direction == "Below fit") %>%
  count(sample_id, sort = TRUE) %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(sample_id, n), y = n)) +
  geom_col(fill = "purple") +
  coord_flip() +
  theme_minimal() +
  ggtitle("Top 20 conditions with outliers BELOW the binomial GLM fit (PHH)") +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0),
    plot.margin = margin(20, 20, 10, 40)
  ) +
  labs(
    x = "Condition (sample_id)",
    y = "Number of outliers below fit"
  )
```

# Outliers per number of modules * number of samples: density of outliers for comparison with other datasets and techniques.

To compare this with RPTEC the number of outliers will be compared relative to the total modules and conditions. Also, to have a more arbitrary number to compare with the other topology-based scoring methods

```{r}
outlier_density <- GRR_glm_fit_df %>%
  filter(outlier, module_number != 0) %>%
  summarise(
    total_outliers = n(),
    total_modules = n_distinct(module_number),
    total_conditions = n_distinct(sample_id),
    density = total_outliers / (total_modules * total_conditions)
  )
```

```{r}
print(outlier_density)
```